\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{titling}
\usepackage{titlesec}
\usepackage{float}
\usepackage[font=small,labelfont=bf,labelsep=period,
  justification=raggedright,singlelinecheck=false]{caption}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=blue!60!black,
  citecolor=black]{hyperref}

\newcommand{\ok}{\textcolor{green!60!black}{OK}}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titlespacing{\section}{0pt}{2.0em}{0.6em}
\titlespacing{\subsection}{0pt}{1.8em}{0.5em}

\setlength{\droptitle}{-4em}
\pretitle{\begin{center}\large\bfseries}
\posttitle{\end{center}\vspace{-1em}}
\preauthor{\begin{center}\normalsize}
\postauthor{\end{center}\vspace{-1em}}
\predate{\begin{center}\normalsize}
\postdate{\end{center}\vspace{-0.5em}}

\title{Decoupled Assessment Pipeline}
\author{Diego Toribio\\[0.3em]{\normalsize March 14, 2026}}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\thispagestyle{empty}

\section{Purpose}

Before launching 86+ batch training runs, we decoupled the
assessment step from training to maximize GPU throughput and
confirmed the post-training pipeline produces correct and complete
artifacts.


\section{Architecture}

\subsection{Problem}

Running assessment episodes inline during training pauses gradient
computation to run 30 episodes per checkpoint. With 10 checkpoints,
this adds 300 episodes per run---20--100\% overhead depending on
episode length. For Rainbow (2 gradient updates per environment
step), this overhead is proportionally more costly.

\subsection{Solution}

\textbf{Training phase} (A100, maximum throughput):
\begin{itemize}[leftmargin=2em, itemsep=0.1em]
  \item Inline assessment disabled
        (\texttt{eval\_every: 999999})
  \item 10 checkpoints saved at uniform intervals
    (\texttt{save\_every: 40000}, \texttt{keep\_last\_n: 10})
  \item No video recording during training
\end{itemize}

\textbf{Assessment phase} (T4 or CPU, cheap):
\begin{itemize}[leftmargin=2em, itemsep=0.1em]
  \item \texttt{reeval\_checkpoints.py} loads each checkpoint,
    runs 30 episodes, saves all artifacts
  \item Produces: \texttt{evaluations.csv},
    \texttt{evaluations.jsonl},
    \texttt{per\_episode\_returns.jsonl},
    \texttt{detailed/*.json}
  \item Optional video recording (\texttt{--record-video}) for
    selected runs only
\end{itemize}

\textbf{Two-tier assessment:} Video recording adds
$\sim$3$\times$ overhead (frame rendering on every step). Fast
assessment without video runs first for all runs. Selective video
recording runs second on chosen checkpoints for thesis figures.


\section{Configuration Changes}

\begin{table}[H]
\centering
\small
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l l l l}
\toprule
\textbf{Parameter} & \textbf{Before} & \textbf{After}
  & \textbf{File} \\
\midrule
\texttt{eval\_every}   & 40000   & 999999
  & \texttt{atari100k\_base.yaml} \\
\texttt{save\_every}   & 100000  & 40000
  & \texttt{atari100k\_base.yaml} \\
\texttt{keep\_last\_n} & 5       & 10
  & \texttt{atari100k\_base.yaml} \\
\bottomrule
\end{tabular}
\caption{Atari-100K base config changes. Rainbow per-game configs
already had matching settings.}
\end{table}


\section{Script Fixes}

Seven issues were identified and fixed:

\medskip
\textbf{1. Missing sticky actions} (critical). The
\texttt{make\_atari\_env()} call did not pass
\texttt{repeat\_action\_probability}. Atari-100K training uses
0.25, but assessment defaulted to 0.0. Scores would have been
incomparable.

\medskip
\textbf{2. NoisyNets mode}. Rainbow with NoisyNets should use
greedy assessment (\texttt{epsilon=0.0}) with noise disabled via
\texttt{model.eval()}. The script used 0.05 for all models.

\medskip
\textbf{3. No video recording}. Added \texttt{--record-video}
flag with best-episode MP4 per checkpoint.

\medskip
\textbf{4. Missing per-episode data}. Added output of
\texttt{per\_episode\_returns.jsonl} and detailed JSON files.

\medskip
\textbf{5. Hardcoded CPU}. Added \texttt{--device} flag for T4
inference.

\medskip
\textbf{6. Missing render mode}. Video recording requires
\texttt{render\_mode="rgb\_array"} passed to the environment.

\medskip
\textbf{7. Missing SPR checkpoint state} (critical).
\texttt{train\_dqn.py} did not pass \texttt{spr\_components} to
checkpoint save calls. Transition model weights, projection heads,
and EMA encoder state were not saved. Without these, transition
model analysis is impossible.


\section{Validation Results}

Tested on Colab T4 (2026-03-14).

\subsection{DQN Boxing}

\begin{table}[H]
\centering
\small
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l r r l}
\toprule
\textbf{Setting} & \textbf{Inline} & \textbf{Reeval}
  & \textbf{Status} \\
\midrule
epsilon        & 0.05           & 0.05
  & \ok \\
sticky actions & 0.0            & 0.0
  & \ok{} (pre-fix run) \\
Score (400K)   & $-3.0 \pm 5.3$ & $-2.1 \pm 5.2$
  & \ok \\
\bottomrule
\end{tabular}
\caption{DQN Boxing: reeval vs.\ inline results.}
\end{table}

\noindent\textit{Timing:} $\sim$1 min/checkpoint without video,
$\sim$3 min with video (30 episodes).

\subsection{Rainbow Crazy Climber}

\begin{table}[H]
\centering
\small
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l r r l}
\toprule
\textbf{Setting} & \textbf{Training} & \textbf{Reeval}
  & \textbf{Status} \\
\midrule
epsilon        & --     & 0.0
  & \ok{} (NoisyNets) \\
sticky actions & 0.25   & 0.25
  & \ok \\
Score (400K)   & 26013  & $24117 \pm 6002$
  & \ok \\
\bottomrule
\end{tabular}
\caption{Rainbow Crazy Climber: reeval vs.\ training output.}
\end{table}

\noindent\textit{Timing:} $\sim$5 min for 30 episodes.


\section{Status}

\begin{itemize}[leftmargin=2em, itemsep=0.3em]
  \item \ok{} Config changes applied
  \item \ok{} Sticky actions correctly read from config
  \item \ok{} NoisyNets greedy mode (epsilon=0.0)
  \item \ok{} Video recording produces valid MP4 files
  \item \ok{} All output formats match inline output
  \item \ok{} DQN and Rainbow scores consistent
  \item \ok{} SPR checkpoint state now saved
\end{itemize}

Pipeline is ready for batch runs. See
\texttt{notes/training-strategy.md} for execution plan and
\texttt{reports/eval.tex} for the figure catalog.

\end{document}
