\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{lineno}
\usepackage{enumitem}
\setlist{itemsep=0.4em, parsep=0pt, before=\vspace{0.5em}, after=\vspace{0.5em}}
\usepackage{xcolor}
\usepackage{titling}
\usepackage{titlesec}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]{hyperref}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titlespacing{\section}{0pt}{1.2em}{0.6em}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titlespacing{\subsection}{0pt}{0.8em}{0.4em}
\titleformat{\subsubsection}{\normalsize\itshape}{\thesubsubsection}{0.5em}{}
\titlespacing{\subsubsection}{0pt}{0.6em}{0.3em}

\setlength{\droptitle}{-4em}
\pretitle{\begin{center}\large\bfseries}
\posttitle{\end{center}\vspace{-1em}}
\preauthor{\begin{center}\normalsize}
\postauthor{\end{center}\vspace{-1em}}
\predate{\begin{center}\normalsize}
\postdate{\end{center}\vspace{-0.5em}}

\title{Thesis Outline}
\author{Diego Toribio\\[0.3em]{\normalsize March 11, 2026}}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\thispagestyle{empty}
\leftlinenumbers
\linenumbers

\section{Introduction}

Deep RL agents need hundreds of millions of interactions to learn, but
data is not always cheap. The Atari-100K benchmark cuts that budget to
100K steps -- roughly two hours of real-time play -- and most standard
agents struggle. Several approaches try to close this gap, and this
thesis focuses on two: self-supervised representation learning (SPR)
and data augmentation.

\subsection{The Data Efficiency Problem}

\begin{itemize}
\item The gap between human and agent data requirements
\item Why data efficiency matters beyond games: sim-to-real, costly
  environments, safety-critical domains
\item Four families of approaches: data augmentation, representation
  learning, world models, algorithmic improvements
\item This thesis focuses on the first two
\end{itemize}

\subsection{The Attribution Problem}

SPR trains the encoder to predict its own future representations, and
it achieved strong results on Atari-100K -- but only on top of Rainbow,
a complex agent with six extensions that each change how the agent
learns. We cannot tell whether SPR's gains came from representation
learning or from Rainbow doing the heavy lifting. Our early results
confirm this matters: SPR on vanilla DQN barely helps in most games.

\begin{itemize}
\item Rainbow's components (distributional values, prioritized replay,
  n-step returns, etc.) confound the result
\item Our early DQN+SPR results: only 1 of 6 games shows a strong
  lift, 4 games are flat or negative
\item Not a failure of SPR -- it shows the base agent matters, and
  nobody has studied when or why
\end{itemize}

\subsection{Research Question and Approach}

\vspace{0.5em}
\begin{quote}
When does self-supervised representation learning improve
data-efficient reinforcement learning, and what determines whether
the learned representations are useful?
\end{quote}
\vspace{0.5em}

We answer this in three parts:

\begin{itemize}
\item \textbf{Isolation} -- Does SPR improve vanilla DQN? How does it
  interact with data augmentation? We test each method alone and in
  combination across six games, three seeds each.
\item \textbf{Capability} -- Does SPR need a capable base agent?
  Rainbow+SPR on the same games provides the contrasting case.
\item \textbf{Interpretability} -- What do the representations
  actually encode, and why do they differ across capability levels?
  Linear probing, latent visualization, and transition model analysis
  on saved checkpoints.
\end{itemize}

\subsection{Contributions}

\begin{itemize}
\item \textbf{Isolation study} -- first evaluation of SPR on vanilla
  DQN, separating representation learning from algorithmic
  improvements
\item \textbf{Interaction study} -- testing augmentation and SPR alone
  and in combination to see whether they are independent or redundant
\item \textbf{Interpretability analysis} -- probing and visualization
  across capability levels, producing a finding about when and why
  self-supervised representations become useful
\item \textbf{Component attribution} (if the investigation yields
  clean results) -- identifying which Rainbow component enables
  useful representations
\item \textbf{Representation-informed modification} -- a targeted
  change to SPR or the training configuration motivated by the
  interpretability findings, testing whether understanding the
  capability gap leads to a concrete improvement
\end{itemize}

\subsection{Thesis Outline}

\begin{itemize}
\item Chapter 2 -- Background: RL, DQN, Rainbow, Atari-100K, SPR,
  augmentation, related work
\item Chapter 3 -- Methods: implementations, experiment design,
  interpretability methods
\item Chapter 4 -- Results: DQN isolation study, capability
  comparison, interpretability findings, cross-game generalization
\item Chapter 5 -- Discussion: interpretation, limitations, broader
  positioning
\item Chapter 6 -- Conclusion: contributions, implications, future work
\end{itemize}

\section{Background}

\noindent Covers the RL foundations, deep learning tools, the DQN agent
and its extensions, the Atari-100K benchmark, and self-supervised
representation learning.

\subsection{Reinforcement Learning}

\noindent Formalizes the problem of learning from interaction and builds
to the algorithms used in DQN. The grid-world serves as a running
example throughout.

\begin{itemize}
\item \textbf{Markov Decision Processes} -- the problem framework:
  agent, environment, states, actions, rewards, and policies.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Transitions, dynamics, Markov property, discount,
    return, episodes, and policies.
  \item Grid-world running example, discounting worked example,
    and three figures.
  \end{itemize}
\item \textbf{Value Functions} -- how to evaluate a policy.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Expected value, state-value v(s), action-value q(s,a),
    and why q is more useful for decision-making.
  \item Value grid and Q-value figures with grid-world examples.
  \end{itemize}
\item \textbf{The Bellman Equation} -- the recursive structure
  of value.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item One-step lookahead and backup operation.
  \item Bellman backup figure and grid-world worked example.
  \end{itemize}
\item \textbf{Optimal Policies} -- what the best policy looks
  like and the Bellman optimality equation.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Optimal value functions, greedy policy, and Bellman
    optimality for both v* and q*.
  \item Grid-world example showing optimal Q-values.
  \end{itemize}
\item \textbf{Q-Learning} -- learning optimal Q-values from
  experience without knowing the dynamics.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item TD updates, bootstrapping, learning rate, exploration
    vs exploitation, and epsilon-greedy.
  \item TD update figure, exploration figure, and worked
    arithmetic.
  \end{itemize}
\item \textbf{Function Approximation} -- why tables fail for large
  state spaces and how parameterized functions replace them.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Parameterized Q(s,a;w), generalization, semi-gradient
    updates, and the deadly triad.
  \item Table-vs-network and semi-gradient figures.
  \end{itemize}
\end{itemize}

\subsection{Deep Learning}

\noindent Builds from the single neuron to full network architectures, covering
training and representations. The reader leaves with everything needed
to understand DQN (2.3) and SPR (2.5).

\begin{itemize}
\item \textbf{Neurons} -- what a single neuron computes.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Weighted sum, bias, activation function, and ReLU.
  \item Neuron diagram, concrete example, and step-by-step
    arithmetic.
  \end{itemize}
\item \textbf{Layers and Depth} -- how neurons compose into
  networks and what depth achieves.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Layers, width, layer types (input, hidden, output), depth,
    and the forward pass.
  \item Why activation functions are essential for depth to be
    useful.
  \end{itemize}
\item \textbf{Network Architecture} -- how the pieces fit together
  into a complete network.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Fully connected layers, MLP, and parameters vs
    hyperparameters.
  \item MLP diagram and parameter count example.
  \end{itemize}
\item \textbf{Convolutional Layers} -- why FC layers fail for
  images and how convolutions fix it.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Kernels, weight sharing, feature maps, and stride.
  \item Convolution operation and stride comparison figures.
  \end{itemize}
\item \textbf{Loss Functions} -- how the network measures its
  errors.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item MSE for value prediction, cross-entropy for classification.
  \item MSE equation with worked example, cross-entropy equation.
  \end{itemize}
\item \textbf{Gradient Descent and Backpropagation} -- how the
  network reduces its errors.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Gradient descent, learning rate, and how backpropagation
    computes gradients layer by layer.
  \item Forward and backward pass figure.
  \end{itemize}
\item \textbf{Optimizers} -- practical tools for training.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item SGD, minibatches, momentum, Adam, and batch normalization.
  \item Connects SGD to the replay buffer sampling from 2.1.
  \end{itemize}
\item \textbf{Representations} -- what the network learns
  internally and why it matters.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Encoder, task head, latent space, what makes a good
    representation, linear probing, and cosine similarity.
  \item Encoder diagram. Bridges to the thesis question about
    representation quality.
  \end{itemize}
\end{itemize}

\subsection{Deep Q-Networks}

\noindent Combines RL from 2.1 with the deep learning tools from 2.2
into a working agent. Covers the full pipeline from raw frames to
Q-values, then extends to Rainbow.

\begin{itemize}
\item \textbf{Preprocessing} -- how raw game frames become network
  input.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Frame max, grayscale, resize, and frame stacking.
  \item Preprocessing pipeline figure and step-by-step visuals.
  \end{itemize}
\item \textbf{The DQN Architecture} -- the network that maps
  frames to Q-values.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Nature CNN: three conv layers, FC head, one Q-value per
    action.
  \item Architecture figure and layer specification table.
  \end{itemize}
\item \textbf{Experience Replay} -- how the agent reuses past
  transitions to break correlation.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Replay buffer, uniform sampling, and data reuse.
  \item Replay buffer diagram.
  \end{itemize}
\item \textbf{Data Augmentation} -- how random spatial shifts
  reduce overfitting when data is limited.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Random shift (pad + crop), why it works in RL, and
    augmentation as implicit regularization.
  \item Random shift and fragile-vs-robust figures.
  \end{itemize}
\item \textbf{Target Networks} -- how a frozen copy of the network
  stabilizes learning.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Moving target problem, frozen copy, periodic updates.
  \item Target network diagram.
  \end{itemize}
\item \textbf{Rainbow Extensions} -- six improvements that address
  different weaknesses in DQN.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Double DQN, prioritized replay, dueling, multi-step,
    distributional (C51), and noisy networks.
  \item Extensions table, dueling and distributional figures.
  \end{itemize}
\item \textbf{The Rainbow Agent} -- how the extensions combine and
  why the encoder is unchanged.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Complementary gains, encoder as the sole unchanged piece.
  \item Rainbow components figure. Bridges to the thesis question
    about base agent capability.
  \end{itemize}
\end{itemize}

\subsection{The Arcade Learning Environment}

\noindent The benchmark for evaluating data-efficient RL. Introduces
the standard and 100K protocols, and the metrics used to compare
agents across games.

\begin{itemize}
\item \textbf{The Standard Protocol} -- the original 200M-frame
  evaluation that established DQN's results.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item ALE interface, frame skip, sticky actions, and game-over
    termination.
  \end{itemize}
\item \textbf{The 100K Protocol} -- 100K steps per game, a
  qualitatively different learning regime.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item 400K frames (~6 minutes), 26-game subset, evaluation
    procedure, and hyperparameter sensitivity.
  \end{itemize}
\item \textbf{Metrics} -- how to compare agents across games with
  different score ranges.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Human-normalized score, IQM, stratified bootstrap
    confidence intervals, and performance profiles.
  \end{itemize}
\end{itemize}

\subsection{Self-Supervised Representation Learning}

\noindent The central topic of the thesis. Introduces self-supervised
learning, the collapse problem, and the specific method (SPR) whose
representations we study.

\begin{itemize}
\item \textbf{Learning Without Labels} -- why the Q-loss alone is
  insufficient and how self-supervised objectives help.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Pretext tasks, temporal structure in RL, predicting in
    latent space, and dual-objective training.
  \item Single-objective and dual-objective architecture figures.
  \end{itemize}
\item \textbf{The Collapse Problem} -- why prediction objectives
  can learn useless representations.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Representational collapse, EMA target encoder, asymmetric
    predictor, and connection to DQN target networks.
  \item Collapse illustration figure.
  \end{itemize}
\item \textbf{Self-Predictive Representations} -- SPR's architecture
  and training procedure.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Transition model, projection and prediction heads,
    multi-step chained predictions, and cosine similarity loss.
  \item SPR architecture and loss figures. Bridges to the thesis
    question: SPR only evaluated on Rainbow.
  \end{itemize}
\end{itemize}

\subsection{Related Work and Positioning}

\noindent Where this thesis sits relative to other data-efficient RL
approaches.

\begin{itemize}
\item \textbf{World Models} -- SimPLe, Dreamer: predictions for
  experience generation.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Differs from SPR: predictions generate experience, not
    shape representations.
  \end{itemize}
\item \textbf{Planning Methods} -- MuZero, EfficientZero: model
  plus search at test time.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Differs from SPR: model used at test time, SPR is
    model-free at test time.
  \end{itemize}
\item \textbf{Training Efficiency} -- BBF, SR-SPR: scaling and
  training improvements.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Both use SPR, so their results cannot be attributed to
    SPR alone.
  \end{itemize}
\item \textbf{Positioning} -- no prior work isolates SPR from its
  base agent or compares representations across capability levels.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Method landscape figure.
  \end{itemize}
\end{itemize}

\section{Methods}

\noindent Covers the agent implementations, the experimental design
that isolates SPR's contribution, and the interpretability tools used
to analyze the learned representations.

\subsection{DQN Implementation}

\noindent Our vanilla DQN, implemented from scratch in PyTorch.

\begin{itemize}
\item \textbf{Architecture} -- Nature CNN encoder over 4 stacked
  grayscale frames.
  \begin{itemize}[label=$\circ$, itemsep=0.2em, parsep=0pt, topsep=0pt, partopsep=0pt]
  \item Three conv layers, FC head, RMSProp optimizer,
    epsilon-greedy with linear decay.
  \item Hyperparameter table and deviations from Mnih 2015.
  \end{itemize}
\end{itemize}

\subsection{Data Augmentation}

DrQ-style random shift applied to sampled batches during training.

\begin{itemize}
\item Pad frames to 92x92, random crop back to 84x84
\item Applied independently to both observations and
  next-observations
\item Configurable via YAML -- same toggle for DQN and Rainbow
  conditions
\end{itemize}

\subsection{Self-Predictive Representations}

SPR is integrated identically on both DQN and Rainbow so the
comparison is controlled.

\begin{itemize}
\item Transition model: two convolutional layers with action
  conditioning via one-hot broadcast
\item EMA target encoder with momentum 0.99
\item Projection and prediction heads (MLP)
\item Cosine similarity loss over K-step predicted representations
\item Combined loss: TD loss + weighted SPR loss
\item Encoder dropout (0.5) for regularization
\end{itemize}

\subsection{Rainbow Implementation}

Rainbow is infrastructure for the capability comparison.

\begin{itemize}
\item Distributional value estimates (C51, 51 atoms)
\item Prioritized experience replay with importance sampling weights
\item Dueling network architecture
\item N-step returns (n=3)
\item Noisy exploration (factorised Gaussian)
\item Double DQN target computation
\item Per-game configs for all six games
\end{itemize}

\subsection{Experiment Design}

We test SPR across two agents of different capability.

\begin{itemize}
\item Part 1: DQN isolation study -- four conditions (baseline,
  +augmentation, +SPR, +both) across 6 games, 3 seeds each
\item Part 2: Rainbow comparison -- Rainbow and Rainbow+SPR on the
  same 6 games, 3 seeds each
\item Game selection: six games spanning dense to moderate reward,
  chosen for diverse interaction patterns in published results
\item Evaluation: 30 episodes at final checkpoint, mean return with
  standard deviation
\item Compute: Google Colab Pro with automated Drive sync
\end{itemize}

\subsection{Interpretability Methods}

Three analyses on saved checkpoints, plus two follow-up
investigations to identify which agent properties enable useful
representations.

\begin{itemize}
\item Linear probing: train classifiers on frozen encoder outputs to
  decode game-relevant features across all conditions
\item Latent space visualization: t-SNE/PCA of encoder outputs,
  comparing structure across DQN and Rainbow conditions
\item Transition model analysis: cosine similarity between predicted
  and actual next-states over K steps
\item Metric correlation: correlate Rainbow training metrics with
  probe accuracy to narrow down which components matter
\item Targeted ablations on the strongest candidates from the
  correlation analysis
\end{itemize}

\subsection{Representation-Informed Modification}

Using the interpretability findings to motivate a targeted change and
test whether it improves performance.

\begin{itemize}
\item What the probes and visualizations reveal about the bottleneck
  on vanilla DQN -- which missing property to address
\item The specific modification: a change to SPR, the training
  configuration, or the addition of a single Rainbow component
\item Experimental setup: same six games, same evaluation protocol,
  controlled comparison against the unmodified conditions
\end{itemize}

\section{Results}

We present results with minimal interpretation -- the Discussion
chapter handles the why. Each section builds on the previous: DQN
conditions first, then Rainbow comparison, then what the
representations look like, then whether findings generalize.

\subsection{DQN Isolation Study}

Four conditions on vanilla DQN tested across six games, three seeds
each.

\begin{itemize}
\item Baseline DQN performance per game
\item Augmentation effect: which games benefit, which collapse
\item SPR effect: modest overall, strong in one game, flat or
  negative in most
\item Combined condition: whether augmentation and SPR together
  outperform either alone
\item Summary table and learning curves with confidence intervals
\end{itemize}

\subsection{Rainbow Capability Comparison}

Rainbow and Rainbow+SPR on the same six games, compared against the
DQN conditions.

\begin{itemize}
\item Rainbow baseline vs vanilla DQN -- where the capable agent
  helps and where it doesn't under 100K steps
\item Rainbow+SPR vs DQN+SPR -- the capability gap and its magnitude
  per game
\item Whether SPR's contribution scales with base agent capability
\end{itemize}

\subsection{Interpretability Analysis}

What the learned representations look like across conditions.

\begin{itemize}
\item Linear probe accuracy: can frozen encoder outputs decode
  game-relevant features? How does this differ across DQN, DQN+SPR,
  Rainbow, Rainbow+SPR?
\item Latent space visualization: t-SNE/PCA comparing representation
  structure across conditions
\item Transition model predictions: how accurately does the model
  predict future states, and does this differ between DQN+SPR and
  Rainbow+SPR?
\item Metric correlations: which Rainbow training metrics track
  representation quality?
\item Targeted ablation results, if conducted
\end{itemize}

\subsection{Cross-Game Generalization}

Whether the findings hold across different types of games.

\begin{itemize}
\item Does the relationship between agent capability and
  representation quality hold across dense, moderate, and sparse
  reward games?
\item Game-specific patterns vs universal findings
\item Whether reward structure predicts where SPR helps
\end{itemize}

\subsection{Representation-Informed Modification}

Whether the interpretability-motivated change improves performance.

\begin{itemize}
\item Performance comparison: modified condition vs the unmodified
  DQN+SPR baseline across six games
\item Whether the modification closes part of the gap between DQN+SPR
  and Rainbow+SPR
\item Probe and visualization comparison: does the modification change
  what the representations encode?
\end{itemize}

% ================================================================
% Chapters 5 and 6 -- to be filled out next
% ================================================================

\section{Discussion}

The Results chapter shows what happened; this chapter explains what it
means. We interpret the capability gap, the interaction between
augmentation and SPR, and what the representations actually encode --
then honestly assess what the findings do and do not establish.

\subsection{The Capability Threshold}

The central finding: SPR barely helps vanilla DQN but is known to
produce large improvements on Rainbow. What does the capable agent
provide that makes self-supervised representations useful?

\begin{itemize}
\item What the DQN vs Rainbow comparison reveals about when
  representation learning works
\item Whether a single Rainbow component explains the gap, or whether
  it is the combination
\item How the metric correlation and ablation results (if available)
  narrow down the mechanism
\end{itemize}

\subsection{Augmentation and Representation Learning}

Augmentation and SPR are both ways to get more out of limited data,
but they work differently -- augmentation regularizes the input space,
SPR shapes the representation. Do they help for the same reasons, or
different ones?

\begin{itemize}
\item Whether augmentation and SPR provide independent or redundant
  improvements
\item Connection to the CURL finding: if augmentation is what matters,
  does adding SPR on top change anything?
\item Per-game variation and what drives it
\end{itemize}

\subsection{What Representations Encode}

The interpretability results tell us what SPR actually learns on each
base agent. This is where the thesis goes beyond performance numbers.

\begin{itemize}
\item What the linear probes reveal about feature encoding across
  conditions
\item Whether SPR produces structured representations only on capable
  agents, or always
\item What transition model accuracy says about where the bottleneck
  is -- representation quality vs downstream policy
\end{itemize}

\subsection{Limitations}

An honest assessment of what the findings do and do not establish.

\begin{itemize}
\item Scale: 6 games, 3 seeds -- enough for patterns, not enough for
  strong statistical claims
\item Algorithm scope: only DQN and Rainbow, no intermediate agents
\item Compute constraints: Colab Pro limits run length and
  parallelism
\item Interpretability scope: limited probe features, qualitative
  t-SNE
\item Augmentation scope: only random shift, no comparison across
  augmentation types
\end{itemize}

\subsection{Broader Context}

How this work connects to the broader ML landscape.

\begin{itemize}
\item Self-supervised pre-training works in NLP and vision -- our
  results are a data point about when it works in RL
\item Honest positioning against published baselines: we are not
  competing with SOTA, we are understanding when representation
  learning helps
\item Implications for practitioners choosing whether to add SPR to
  their agent
\end{itemize}

\section{Conclusion}

Short chapter. Restates the research question, summarizes what we
found, and looks forward.

\subsection{Summary of Contributions}

Restate the research question and summarize each contribution with
actual findings.

\begin{itemize}
\item Isolation: what SPR contributes on vanilla DQN without
  Rainbow's confounds
\item Interaction: whether augmentation and SPR are independent or
  redundant on a simple agent
\item Interpretability: what the representations encode and why
  capability matters
\item Component attribution (if applicable): which Rainbow property
  enables useful representations
\item Representation-informed modification: whether understanding the
  capability gap leads to a concrete improvement on the simple agent
\end{itemize}

\subsection{Implications}

What the findings mean for people building RL agents and for
researchers studying representation learning.

\begin{itemize}
\item For practitioners: when SPR is worth adding, whether
  augmentation alone is sufficient on simple agents, what to expect
  under tight data budgets
\item For researchers: what interpretability reveals about
  self-supervised objectives in RL, and the isolation study as a
  reference point for future work on SPR
\end{itemize}

\subsection{Future Work}

Concrete directions motivated by what we found and what we could not
cover.

\begin{itemize}
\item Expanded game set: the full 26-game Atari-100K benchmark
\item Intermediate agents: testing DQN with individual Rainbow
  components to map the capability threshold more precisely
\item Contrastive comparison: CURL vs SPR on the same conditions
\item Cross-game transfer: whether representations learned on one
  game help on another
\item Scaling: larger encoders, more seeds, cloud compute
\end{itemize}

\end{document}
