\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{titling}
\usepackage{titlesec}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=blue!60!black,citecolor=black]{hyperref}

% Inline references as clickable [N] links (no bibliography section)
\newcommand{\citelink}[2]{\href{#1}{\small[#2]}}

\renewcommand\linenumberfont{\normalfont\scriptsize\sffamily\color[rgb]{0.5,0.5,0.5}}
\leftlinenumbers
\linenumbers

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titlespacing{\section}{0pt}{2.0em}{0.6em}
\titlespacing{\subsection}{0pt}{1.8em}{0.5em}

\setlength{\droptitle}{-4em}
\pretitle{\begin{center}\large\bfseries}
\posttitle{\end{center}\vspace{-1em}}
\preauthor{\begin{center}\normalsize}
\postauthor{\end{center}\vspace{-1em}}
\predate{\begin{center}\normalsize}
\postdate{\end{center}\vspace{-0.5em}}

\title{Results}
\author{Diego Toribio\\[0.3em]{\normalsize March 10, 2026}}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\thispagestyle{empty}

\noindent{\small This document is a working snapshot of experimental
results. Baseline runs are in progress; augmentation and SPR conditions
have not yet been run.}

\section{Atari-100K Benchmark}

Atari-100K \citelink{https://arxiv.org/abs/1903.00374}{1} is a
data-efficiency benchmark for reinforcement learning. The agent gets
just 100K environment steps per game -- roughly two hours of real-time
play. That is two orders of magnitude less experience than the standard
protocol used to train the original DQN
\citelink{https://www.nature.com/articles/nature14236}{2}, so the
question becomes: can an agent learn anything useful with so little
data?

\medskip
An \textit{environment step} is one decision point: the agent sees a
game screen, picks an action, and receives a reward. Under the standard
frame skip of 4, each decision is held for 4 consecutive video frames,
so 100K steps correspond to 400K rendered frames.

\medskip
SPR \citelink{https://arxiv.org/abs/2007.05929}{3} adds a
self-supervised objective that trains the agent to predict its own
future internal states, encouraging richer representations even when
reward is scarce. However, SPR was only evaluated on top of Rainbow
\citelink{https://arxiv.org/abs/1710.02298}{4} -- a complex agent with
many components -- so it is unclear whether the representation learning
is what helped or whether Rainbow did the heavy lifting. We isolate SPR
on vanilla DQN to answer that question, using \textit{data
augmentation} (small random shifts to each game frame) as a second
factor. By testing each method alone and in combination across six
games, we can isolate their individual and combined effects.

\section{Experimental Design}

\subsection{What We Compare}

We start with a vanilla DQN baseline and progressively add augmentation
and SPR. Each condition is run across six games with three seeds
(72 runs total).

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{12pt}
\begin{tabular}{l l}
\toprule
\textbf{Condition} & \textbf{What changes} \\
\midrule
Baseline  & Vanilla DQN, no modifications \\
+ Aug     & Add random-shift data augmentation \\
+ SPR     & Add self-predictive representation learning \\
+ Both    & Augmentation and SPR combined \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 1.} The four experimental conditions.
Each builds on the same DQN architecture and hyperparameters.\par}
\end{table}

\subsection{Game Selection}

We selected six games to cover a range of expected interactions between
augmentation and SPR, based on published results in Schwarzer et al.\
\citelink{https://arxiv.org/abs/2007.05929}{3}:

\begin{itemize}[itemsep=2pt]
  \item \textbf{Crazy Climber} -- SPR dominates; augmentation adds little
  \item \textbf{Boxing} -- both SPR and augmentation contribute
    independently
  \item \textbf{Kangaroo} -- strong synergy; both components needed
  \item \textbf{Road Runner} -- SPR large lift; augmentation moderate
    boost
  \item \textbf{Frostbite} -- SPR dominant ($6\times$ over baseline);
    augmentation adds only 10\%
  \item \textbf{Up N Down} -- augmentation dominant; boosts SPR by
    $4\times$
\end{itemize}

\subsection{Training Setup}

All conditions use the Nature CNN architecture
\citelink{https://www.nature.com/articles/nature14236}{2} as the shared
encoder and are trained with the same configuration. Augmentation and
SPR add their own components on top of this base. Results report mean
raw episode return ($\pm$ std) across three seeds.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.2}
\setlength{\tabcolsep}{12pt}
\begin{tabular}{l l}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Optimizer       & RMSProp (lr = 0.00025) \\
Batch size      & 32 \\
Exploration     & $\epsilon$: 1.0 $\to$ 0.1 \\
Replay buffer   & 100K capacity, 2K warmup \\
Target network  & Hard copy every 2K steps \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 2.} Training configuration shared
across all four conditions.\par}
\end{table}

\noindent The agent explores by choosing random actions with probability
$\epsilon$, which decays from 1.0 (fully random) to 0.1 over the first
half of training. The replay buffer stores past transitions so the agent
can retrain on earlier experience; training begins after 2K transitions
have been collected. The target network is a frozen copy of the main
network used to compute stable learning targets, refreshed every 2K
steps.

\section{Results}

\subsection{Baseline Validation (seed 42)}

Initial single-seed runs on three games to validate the training
pipeline. These runs used clipped rewards in evaluation (see
Appendix~\ref{app:clipping}), so scores are not comparable to published
baselines.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{l r r r r r}
\toprule
\textbf{Game}
  & \textbf{Final Eval}
  & \textbf{Best Eval}
  & \textbf{Episodes}
  & \textbf{Avg Length}
  & \textbf{Time} \\
\midrule
Crazy Climber  & $152.7 \pm 37.4$   & $162.5$ (320K)  & 182 & $\sim$2200  & 43 min \\
Kung Fu Master & $1.4 \pm 1.6$      & $3.7$ (320K)    & 458 & $\sim$870   & 16 min \\
Boxing         & $-19.7 \pm 7.1$    & $-0.2$ (40K)    & 56  & $\sim$7100  & 29 min \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 3.} Vanilla DQN baseline at 100K env
steps in clipped reward space. 30 eval episodes per checkpoint,
eval epsilon 0.05.\par}
\end{table}

\noindent Only Crazy Climber shows clear learning (monotonic improvement
with high variance). Kung Fu Master remains flat and Boxing converges to
a losing policy. This is expected -- vanilla DQN struggles at 100K
steps, which is precisely why augmentation and SPR are needed.

\subsection{Baseline (seed 42, raw scores)}

Six-game baseline with the eval clipping fix applied. Single seed (42)
per game; 30 evaluation episodes per checkpoint.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{l r r r r}
\toprule
\textbf{Game}
  & \textbf{Final Eval (400K)}
  & \textbf{Best Eval}
  & \textbf{Random}
  & \textbf{Human} \\
\midrule
Crazy Climber  & $3540 \pm 754$    & $17460$ (320K) & 10780 & 35829 \\
Road Runner    & $1440 \pm 859$    & $1633$ (160K)  & 11500 & 7845  \\
Boxing         & $-3.0 \pm 5.3$    & $-3.0$ (400K)  & 0.1   & 12.1  \\
Kangaroo       & $26.7 \pm 68.0$   & $293$ (120K)   & 52    & 3035  \\
Frostbite      & $147 \pm 45$      & $168$ (360K)   & 65    & 4335  \\
Up N Down      & $141 \pm 155$     & $1312$ (120K)  & 533   & 11693 \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 4.} Vanilla DQN baseline on the six
selected games with raw (unclipped) evaluation scores. Random and human
scores from Schwarzer et al.\ \citelink{https://arxiv.org/abs/2007.05929}{3}.\par}
\end{table}

\subsection{Augmentation}

DrQ-style random shift augmentation
\citelink{https://arxiv.org/abs/2004.14990}{5}: pad each 84$\times$84
frame by 4 pixels on each side (zero-fill), then randomly crop back to
84$\times$84. Applied independently to states and next-states at each
training step.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{12pt}
\begin{tabular}{l r r}
\toprule
\textbf{Game} & \textbf{Baseline} & \textbf{+ Aug} \\
\midrule
Crazy Climber  & $3540 \pm 754$   & $110 \pm 40$ \\
Road Runner    & $1440 \pm 859$   & $3.3 \pm 18.0$ \\
Boxing         & $-3.0 \pm 5.3$   & $-18.2 \pm 4.0$ \\
Kangaroo       & $26.7 \pm 68.0$  & $460 \pm 128$ \\
Frostbite      & $147 \pm 45$     & $0.0 \pm 0.0$ \\
Up N Down      & $141 \pm 155$    & $981 \pm 671$ \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 5.} Effect of random-shift data
augmentation compared to the vanilla DQN baseline. Seed 42, 30 eval
episodes at 400K frames.\par}
\end{table}

\subsection{SPR}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{12pt}
\begin{tabular}{l r r}
\toprule
\textbf{Game} & \textbf{Baseline} & \textbf{+ SPR} \\
\midrule
Crazy Climber  & --  & -- \\
Boxing         & --  & -- \\
Kangaroo       & --  & -- \\
Road Runner    & --  & -- \\
Frostbite      & --  & -- \\
Up N Down      & --  & -- \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 6.} Effect of self-predictive
representations compared to the vanilla DQN baseline.\par}
\end{table}

\subsection{Combined}

\noindent The combined condition (+Both) uses the same EMA momentum
(0.99) and dropout (0.5) as the SPR-only condition, with augmentation
as the sole addition. This isolates augmentation's contribution on top
of SPR rather than conflating it with changes to regularization
strategy.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{12pt}
\begin{tabular}{l r r r}
\toprule
\textbf{Game} & \textbf{Baseline} & \textbf{+ Aug} & \textbf{+ Both} \\
\midrule
Crazy Climber  & --  & --  & -- \\
Boxing         & --  & --  & -- \\
Kangaroo       & --  & --  & -- \\
Road Runner    & --  & --  & -- \\
Frostbite      & --  & --  & -- \\
Up N Down      & --  & --  & -- \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 7.} Effect of combining augmentation
and SPR, compared to augmentation alone and the baseline.\par}
\end{table}

\subsection{Summary}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{14pt}
\begin{tabular}{l rrrr}
\toprule
\textbf{Game}
  & \textbf{Baseline}
  & \textbf{+ Aug}
  & \textbf{+ SPR}
  & \textbf{+ Both} \\
\midrule
Crazy Climber  & --  & --  & --  & --  \\
Boxing         & --  & --  & --  & --  \\
Kangaroo       & --  & --  & --  & --  \\
Road Runner    & --  & --  & --  & --  \\
Frostbite      & --  & --  & --  & --  \\
Up N Down      & --  & --  & --  & --  \\
\midrule
Mean           & --  & --  & --  & --  \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 8.} All conditions. Mean raw episode
return ($\pm$ std) across 3 seeds per condition.\par}
\end{table}

\clearpage

\appendix
\section*{Appendix}
\addcontentsline{toc}{section}{Appendix}

\section{Known Issue: Reward Clipping in Evaluation}
\label{app:clipping}

The initial baseline validation (Table 3) reported returns in
\textbf{clipped reward space} ($\{-1, 0, +1\}$ per step) because the
evaluation environment inherited the training reward clipper. Published
baselines
\citelink{https://arxiv.org/abs/2007.05929}{3}%
\citelink{https://arxiv.org/abs/2004.14990}{5}
report \textbf{raw Atari scores}. The two scales are not comparable --
for example, the random agent scores 10,780 raw on Crazy Climber while
our clipped-reward agent reports 153. This was fixed for all subsequent
runs (Table 3 onward) by disabling reward clipping in the evaluation
environment.

\section{Early Baseline Observations}
\label{app:observations}

Detailed per-game analysis from the initial three-game validation run
(Table 3):

\begin{enumerate}[nosep]
  \item \textbf{Crazy Climber shows clear learning.} Monotonic
    improvement through 320K with variance-driven dips. Loss converged
    from 0.87 to 0.055 (moving average). Viable baseline anchor for the
    comparative study.

  \item \textbf{Kung Fu Master shows no learning.} Eval never exceeds
    3.7 (clipped). 458 episodes in 400K frames indicates the agent
    dies quickly ($\sim$870 frames per episode). Loss converged normally
    (2.49$\to$0.024), suggesting the network fits Q-values but they do
    not translate to useful behavior within the 100K budget.

  \item \textbf{Boxing learned a losing policy.} Eval oscillates
    wildly ($-0.2$ to $-42.5$), with the agent actively losing by large
    margins in late training. Only 56 episodes in 400K frames
    ($\sim$7100 frames each) meant the agent saw fewer than 28 episodes
    before epsilon reached 0.1 at 200K frames.

  \item \textbf{Epsilon decay may be too aggressive for long-episode
    games.} By 200K frames, Crazy Climber had seen $\sim$90 episodes
    (sufficient), Kung Fu Master $\sim$230 (sufficient but short-lived),
    and Boxing only $\sim$28 (insufficient for an 18-action game).
\end{enumerate}

\section{Game Selection Reference}
\label{app:games}

The 26 Atari-100K benchmark games grouped by reward density -- how
often the agent receives a non-zero score during play. In dense games
the agent scores on most actions; in sparse games, long stretches pass
with no reward. Of the 26 games, 7 are dense, 12 moderate, and
4 sparse. Categorization based on Schwarzer et al.\
\citelink{https://arxiv.org/abs/2007.05929}{3}, Table 4.

\vspace{0.3em}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.15}
\setlength{\tabcolsep}{14pt}
\begin{tabular}{l l l}
\toprule
\textbf{Dense} & \textbf{Moderate} & \textbf{Sparse} \\
\midrule
Road Runner     & Seaquest    & Pong        \\
Crazy Climber   & Frostbite   & Freeway     \\
Kung Fu Master  & Hero        & Private Eye \\
Up N Down       & Alien       & Bank Heist  \\
Krull           & Asterix     &             \\
Demon Attack    & Ms.\ Pac-Man &            \\
Chopper Command & BattleZone  &             \\
                & Kangaroo    &             \\
                & Breakout    &             \\
                & Jamesbond   &             \\
                & Boxing      &             \\
                & Gopher      &             \\
\bottomrule
\end{tabular}

\vspace{1.2em}
{\raggedright\small \textbf{Table 9.} Atari-100K benchmark games grouped by
how frequently the agent receives non-zero reward. Dense games yield
reward on most steps; sparse games have long stretches of zero reward.\par}
\end{table}

\end{document}
