\chapter{Background and Related Work}
\label{ch:background}

\section{Reinforcement Learning Foundations}
\label{sec:rl-foundations}

This section introduces the core concepts that underpin modern
deep reinforcement learning. We start with the Markov decision
process framework that defines the problem, then develop value
functions and Q-learning as the solution approach, and finally
show why function approximation is needed to scale these ideas
to high-dimensional environments.

\subsection{Markov Decision Processes}
\label{subsec:mdp}

In reinforcement learning, an agent learns by interacting with its
environment -- observing situations, taking actions, and receiving
feedback. The \emph{Markov decision process} (MDP) is the
mathematical framework that makes this interaction
precise~\citep{sutton2018reinforcement}, and nearly every
reinforcement learning algorithm -- from tabular methods to deep
neural network approaches -- is built on it.

\smallskip
To make the framework concrete, consider the grid-world in
\cref{fig:grid-world}: an agent starts in one cell and must
navigate to a goal by choosing to move up, down, left, or right
at each step. Each cell the agent can occupy is a \emph{state},
each movement direction is an \emph{action}, and the numerical
feedback it receives after moving -- such as $+1$ for reaching the
goal or $-1$ for reaching a penalty -- is a \emph{reward}. The
grid-world is deliberately simple, but every MDP, no matter how
complex, has exactly these three ingredients: states, actions, and
rewards.

\input{figures/grid-world}

\smallskip
More generally, an MDP describes a loop between an agent and an
environment (\cref{fig:mdp-loop}). At each discrete time step
$t = 0, 1, 2, \ldots$, the agent observes a state
$S_t \in \mathcal{S}$, chooses an action
$A_t \in \mathcal{A}(S_t)$, and receives a scalar reward
$R_{t+1} \in \mathbb{R}$ as the environment transitions to a new
state $S_{t+1}$. In the grid-world, $S_t$ is the agent's current
cell, $A_t$ is the chosen direction, and $R_{t+1}$ is the score
received after the move. One complete cycle of this loop -- the
state the agent was in, the action it chose, the reward it received,
and the state it landed in -- is called a \emph{transition} and is
written as the tuple $(S_t, A_t, R_{t+1}, S_{t+1})$.

\input{figures/mdp-loop}

\smallskip
These components -- states, actions, transitions, and how much
the agent cares about future rewards -- are collected into a
single mathematical object: the MDP tuple
$(\mathcal{S},\allowbreak \mathcal{A},\allowbreak p,\allowbreak \gamma)$, where $\mathcal{S}$
is the set of states, $\mathcal{A}$ the set of actions, $p$ a
dynamics function, and $\gamma$ a discount factor. The dynamics
function captures both the transition probabilities and the
reward structure -- it gives the probability of each possible
next state and reward, given the current state and action:
%
\begin{equation}
\label{eq:mdp-dynamics}
p(s', r \mid s, a)
  \;\doteq\;
  \Pr\{S_t = s',\, R_t = r \mid S_{t-1} = s,\, A_{t-1} = a\}.
\end{equation}
%
In words: given that the agent is in state~$s$ and takes
action~$a$, the function~$p$ gives the probability of ending up
in state~$s'$ with reward~$r$. It is a conditional probability
distribution -- for any state--action pair $(s, a)$, the values
$p(s', r \mid s, a)$ sum to one over all possible next states
$s'$ and rewards $r$.

\smallskip
In the grid-world, moving right from a given cell always takes
the agent to the cell on its right -- the dynamics are
deterministic. But many environments are not: wind can push a
robot off course, a slippery surface can send it sliding past
its target, or an opponent's response can change the outcome.
The dynamics function $p$ is a probability distribution rather
than a deterministic mapping precisely because it must handle
these uncertain outcomes.

\smallskip
The key assumption underlying this framework is the \emph{Markov
property}: the probability of reaching any next state and reward
depends only on the current state and action -- not on the history
of states and actions that came before. In the grid-world, knowing
the agent is in cell~$s_{21}$ is enough to determine what happens
when it moves up; it does not matter whether the agent arrived from
below, from the left, or by a longer winding path.

\smallskip
This is what makes the problem tractable. If the outcome depended
on the agent's entire history -- every state it had visited and
every action it had taken -- the agent would need to track an
ever-growing record of past experience. The Markov property says
that the current state is a sufficient summary: once we know where
the agent is now, the past is irrelevant for predicting the future.

\smallskip
\cref{fig:mdp-views} illustrates several transitions from the
grid-world: each arrow is one entry of the dynamics function, with
the action label, destination state, and reward annotation
corresponding to $a$, $s'$, and $r$.

\input{figures/mdp-graph}

\smallskip
A note on notation: the subscripts in \cref{eq:mdp-dynamics} shift
by one relative to the prose above -- the agent in state $S_{t-1}$
takes action $A_{t-1}$ and then receives $R_t$ and $S_t$, which is
the same relationship as $S_t, A_t \to R_{t+1}, S_{t+1}$ with
different index labels. Both conventions appear in the literature;
the distinction is purely notational.

\smallskip
The agent's goal is to collect as much reward as possible over
time -- but rewards at different times are not equally valuable.
A reward received now is worth more than the same reward received
far in the future, both because the future is uncertain and because
earlier rewards can be built upon sooner. The \emph{return} makes
this precise -- it is the cumulative discounted reward from time
step $t$ onward:
%
\begin{equation}
\label{eq:return}
G_t \;\doteq\; R_{t+1} + \gamma\, R_{t+2} + \gamma^2 R_{t+3} + \cdots
  = \sum_{k=0}^{\infty} \gamma^k R_{t+k+1},
\end{equation}
%
Each term $\gamma^k R_{t+k+1}$ weights a future reward by how
far away it is -- rewards received sooner count more, and very
distant rewards fade toward zero.

\smallskip
To see this concretely, suppose the agent receives rewards
$0, 0, +1$ over three steps with $\gamma = 0.9$. The return
from the first step is $G_0 = 0 + 0.9 \cdot 0 + 0.81 \cdot 1
= 0.81$ -- the distant reward of $+1$ is worth only $0.81$
today because it is discounted twice.

\smallskip
The parameter $\gamma \in [0, 1)$ is the \emph{discount rate}. When
$\gamma = 0$, the agent cares only about the very next reward and
ignores everything after -- a myopic strategy. As $\gamma$
approaches 1, the agent plans further ahead, weighing future
rewards more heavily.

\smallskip
Discounting also serves a mathematical purpose: because
$\gamma < 1$, the infinite sum in \cref{eq:return} converges as
long as individual rewards are bounded. Without discounting
($\gamma = 1$), the return could be infinite, making it impossible
to compare policies meaningfully.
\cref{fig:discounting} shows how this plays out: with a low
discount rate, a nearby small reward is preferred over a distant
large one; with a high discount rate, the agent walks further to
collect the larger reward.

\input{figures/discounting}

\smallskip
In many problems, the interaction breaks naturally into
\emph{episodes} -- finite sequences that end in a terminal state
and then reset. The grid-world is episodic: the agent starts in
cell $s_{00}$, takes a sequence of moves, and the episode ends
when it reaches the goal or the penalty. Most reinforcement
learning benchmark environments are episodic, so in practice the
return reduces to a finite sum over the steps within each episode.

\smallskip
Given an MDP, the agent needs a rule for choosing actions. This
rule is called a \emph{policy}, written $\pi(a \mid s)$, which
gives the probability of selecting action~$a$ in state~$s$. A
deterministic policy always picks the same action in a given state,
while a stochastic policy samples from a distribution over actions.

\smallskip
In the grid-world, a policy might say: ``always go right in
$s_{00}$, always go up in $s_{10}$, always go up in $s_{20}$.''
This deterministic policy commits to a fixed direction in each
cell, tracing a specific path through the grid. A different
policy -- choosing different directions in some cells -- would
trace a different path and potentially collect different rewards.

\smallskip
The central question of reinforcement learning is how to find a
policy that maximizes the expected return -- and answering that
question requires a way to evaluate how good a given policy is,
which leads to value functions.

\subsection{Value Functions}
\label{subsec:value-functions}

A policy tells the agent what to do, but not how well it will do.
To compare policies -- and ultimately find the best one -- the
agent needs a way to assign a numerical score to each state,
capturing how much reward it can expect from that point onward.

\smallskip
The \emph{state-value function} $v_\pi(s)$ answers exactly this
question: it gives the expected return when starting in state~$s$
and following policy~$\pi$ thereafter:
%
\begin{equation}
\label{eq:v-pi}
v_\pi(s) \;\doteq\; \mathbb{E}_\pi\!\big[G_t \mid S_t = s\big].
\end{equation}
%
The subscript~$\pi$ on the expectation means that the agent
follows policy~$\pi$ at every step -- the average is taken over
all the randomness in the trajectory (action choices and
transitions) that this policy produces.

\smallskip
In the grid-world, $v_\pi(s)$ is the average total reward the
agent collects from cell~$s$ onward if it follows policy~$\pi$.
Cells near the goal have high values because the agent reaches
the $+1$ reward quickly; cells near the penalty have low values
because the policy is likely to lead there.

\smallskip
\cref{fig:value-grid} makes this concrete: panel~(a) shows the
value of each cell under a policy that navigates toward the goal,
while panel~(b) shows the policy itself. The values form a
gradient -- highest near the goal, decreasing with distance.

\input{figures/value-grid}

\smallskip
The state-value function tells us how good a state is, but it
does not tell us which action to take. For that, we need the
\emph{action-value function} $q_\pi(s, a)$, which gives the
expected return for taking action~$a$ in state~$s$ and then
following~$\pi$:
%
\begin{equation}
\label{eq:q-pi}
q_\pi(s, a) \;\doteq\; \mathbb{E}_\pi\!\big[G_t \mid S_t = s,\, A_t = a\big].
\end{equation}
%
The difference is subtle but important: $v_\pi$ evaluates
\emph{states}, while $q_\pi$ evaluates \emph{actions in states}.

\smallskip
In the grid-world, consider the agent in cell~$s_{21}$.
Moving up leads toward the $+1$ goal, so
$q_\pi(s_{21}, \text{up})$ is high. Moving right leads directly
to the $-1$ penalty in $s_{31}$, so
$q_\pi(s_{21}, \text{right})$ is low. The state value $v_\pi$
assigns a single number to~$s_{21}$, but the action values reveal
that the actions available there are dramatically different.
\Cref{fig:q-values} shows this: each arrow from~$s_{21}$ carries
a different Q-value, making the best action immediately apparent.

\input{figures/q-values}

\smallskip
This is why $q_\pi$ is more directly useful for decision-making
than $v_\pi$: comparing Q-values across actions tells the agent
which action to take, without needing a model of the environment.

\smallskip
Both value functions are defined as expectations over the entire
future -- averages over every possible trajectory from the current
state onward. Computing them from these definitions directly would
require enumerating all such trajectories, which quickly becomes
impractical. The next section shows that the Markov property provides
a recursive shortcut.

\subsection{The Bellman Equation}
\label{subsec:bellman-equation}

Computing $v_\pi$ from its definition means averaging the return over
every possible trajectory from state~$s$ -- every sequence of
actions, transitions, and rewards stretching to the end of the
episode. Even for the grid-world, the number of trajectories grows
exponentially with the episode length. For larger problems, this
direct approach is hopeless.

\smallskip
The Markov property from \cref{subsec:mdp} provides a way out.
Because the future depends only on the current state, the value of a
state can be decomposed into two pieces: the reward from the next
step, and the discounted value of wherever the agent ends up. For
$v_\pi$, this gives the \emph{Bellman expectation equation}:
%
\begin{equation}
\label{eq:bellman-expectation}
v_\pi(s) = \sum_{a} \pi(a \mid s) \sum_{s',\, r}
  p(s', r \mid s, a)\,\big[r + \gamma\, v_\pi(s')\big].
\end{equation}
%
In words: the value of state~$s$ is the average, over all actions the
policy might take, of the immediate reward plus the discounted value
of the next state.

\smallskip
The equation has three layers, each handling a different source of
uncertainty. The outer sum weights actions by $\pi(a \mid s)$ -- the
probability the policy assigns to each action. The inner sum weights
outcomes by $p(s', r \mid s, a)$ -- the environment's transition
probabilities. And the bracketed term $r + \gamma\, v_\pi(s')$ is the
one-step payoff: the reward received now, plus the discounted value of
the successor state. \cref{fig:bellman-backup} shows this structure
as a tree: from state~$s$, the policy selects an action, the
environment responds with a reward and next state, and the payoff at
each leaf is the immediate reward plus the discounted future value.

\input{figures/bellman-backup}

\smallskip
To see this concretely, consider cell~$s_{21}$ in the grid-world.
Suppose the policy says ``go up'' with certainty, and moving up from
$s_{21}$ always leads to $s_{22}$ with reward $r = 0$ (the transition
is deterministic). Then both sums collapse to a single term:
%
\[
v_\pi(s_{21}) = 0 + 0.9 \times v_\pi(s_{22}).
\]
%
If $v_\pi(s_{22}) = 1.0$, then $v_\pi(s_{21}) = 0.90$ -- matching
the values shown in \cref{fig:value-grid}. The value of~$s_{21}$ is
entirely determined by the value of its neighbor; no need to trace the
trajectory all the way to the goal.

\smallskip
When the policy is stochastic -- assigning positive probability to
more than one action -- the value becomes a weighted average across
all possible actions, each weighted by the probability the policy
assigns to it. And when transitions are stochastic, there is an
additional average over the possible next states the environment might
produce. The equation handles both cases through its nested sums.

\smallskip
This equation might seem circular -- the value of a state is defined
in terms of the values of other states, which are themselves defined
the same way. But it is not circular; it is recursive. The chain of
dependencies terminates at terminal states, where the value is zero
because the episode is over and no more reward can be collected. From
there, values propagate backward: the cell next to the goal derives
its value from the goal, the cell two steps away derives its value
from the cell one step away, and so on through the grid.

\smallskip
The power of the Bellman equation is that it reduces a calculation
over the entire future -- the sum of all discounted rewards along
every possible trajectory -- to a one-step relationship between
neighboring states. Instead of tracing every path to the end of the
episode, we only need to look one step ahead. This recursion is what
makes reinforcement learning computationally tractable.

\subsection{Optimal Policies}
\label{subsec:optimal-policies}

The Bellman equation tells us how to evaluate any given policy -- but
the agent does not just want to evaluate its current strategy. It
wants to find the best one.

\smallskip
A policy $\pi^*$ is \emph{optimal} if its expected return is at least
as high as that of every other policy, in every state:
$v_{\pi^*}(s) \geq v_\pi(s)$ for all states~$s$ and all
policies~$\pi$~\citep{sutton2018reinforcement}. At least one optimal
policy always exists.

\smallskip
There may be several optimal policies -- for instance, two paths
through the grid-world might yield the same total reward. But all
optimal policies share the same value functions, which we denote
$v_*$ and $q_*$:
%
\begin{align}
\label{eq:v-star}
v_*(s) &\;\doteq\; \max_\pi\, v_\pi(s), \\[4pt]
\label{eq:q-star}
q_*(s, a) &\;\doteq\; \max_\pi\, q_\pi(s, a).
\end{align}
%
In words, $v_*(s)$ is the highest value achievable in state~$s$ under
any policy, and $q_*(s, a)$ is the highest value achievable by taking
action~$a$ in state~$s$ and then acting optimally thereafter.

\smallskip
The optimal value functions satisfy their own recursive relationship,
the \emph{Bellman optimality equation}. For $q_*$:
%
\begin{equation}
\label{eq:bellman-optimality-q}
q_*(s, a) = \sum_{s',\, r} p(s', r \mid s, a)
  \Big[r + \gamma \max_{a'} q_*(s', a')\Big].
\end{equation}
%
The key difference from the Bellman expectation equation
(\cref{eq:bellman-expectation}) is that the average over the policy's
action choices is replaced by a maximum: instead of asking ``what
return does the current policy achieve,'' we ask ``what is the best
possible return if we act optimally from here.''

\smallskip
If $q_*$ were known, acting optimally would be trivial: in every
state, the agent simply picks the action with the highest $q$-value,
$\pi^*(s) = \arg\max_a q_*(s, a)$. No model of the environment is
needed and no planning is required -- the optimal action is a direct
lookup. This is why much of reinforcement learning, including the
Q-learning algorithm in the next section and the deep Q-networks that
build on it, focuses on estimating $q_*$ rather than $v_*$.

\smallskip
There is a catch, however. The Bellman optimality equation is a
system of $|\mathcal{S}| \times |\mathcal{A}|$ simultaneous
equations, one for every state--action pair. Solving it requires
knowing the dynamics~$p$ and enumerating every state -- both of which
are impractical for any problem with a large state space or unknown
dynamics. The next two sections address these obstacles: Q-learning
estimates $q_*$ from experience without knowing the dynamics, and
function approximation replaces the explicit table of $q$-values with
a neural network that can generalize across states.

\subsection{Q-Learning}
\label{subsec:q-learning}

The Bellman optimality equation tells us what $q_*$ must satisfy,
but solving it requires knowing the dynamics~$p$. In most problems
of interest, the agent does not have this information -- it cannot
look up what will happen; it can only try an action and observe the
result. Q-learning~\citep{watkins1992q} addresses exactly this
setting: the agent learns $q_*$ directly from experience, one
transition at a time, without ever knowing $p$.

\smallskip
The key idea comes from \emph{temporal-difference} (TD)
learning~\citep{sutton1988learning}: rather than waiting until
the end of an episode to see how things turned out, the agent
updates its estimate after every single step. It observes the
immediate reward, peeks at its current estimate of the next
state's value, and uses these together as a learning target.
This peeking is called \emph{bootstrapping} -- the agent learns
from its own estimates rather than from complete outcomes. The
estimates may be wrong at first, but because each update
incorporates a real observed reward, the errors shrink over time.

\smallskip
The Q-learning update rule puts this idea into practice. After
the agent takes action~$A_t$ in state~$S_t$, observes
reward~$R_{t+1}$, and arrives in state~$S_{t+1}$, it updates
its estimate of $q_*(S_t, A_t)$:
%
\begin{equation}
\label{eq:q-learning}
Q(S_t, A_t) \;\leftarrow\; Q(S_t, A_t)
  + \alpha \Big[R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a)
  - Q(S_t, A_t)\Big].
\end{equation}
%
$Q(S_t, A_t)$ is the agent's current estimate for this
state--action pair. The expression
$R_{t+1} + \gamma \max_a Q(S_{t+1}, a)$ is the \emph{TD
target} -- the reward the agent actually received, plus its best
estimate of the value from the next state onward. The difference
between the target and the current estimate is the \emph{TD
error}: how far off the agent's prediction was. The learning rate
$\alpha \in (0, 1]$ controls how far the estimate moves toward
the target on each step -- large $\alpha$ moves quickly but may
overshoot; small $\alpha$ is more stable but learns slowly.

\smallskip
To see this in the grid-world, suppose the agent is in
cell~$s_{21}$ and moves up, receiving reward $r = 0$ and
arriving in cell~$s_{22}$. If the agent currently estimates
$Q(s_{21}, \text{up}) = 0.5$ and $\max_a Q(s_{22}, a) = 1.0$,
then with $\gamma = 0.9$ and $\alpha = 0.1$:
%
\begin{align*}
\text{target} &= 0 + 0.9 \times 1.0 = 0.9, \\
\text{TD error} &= 0.9 - 0.5 = 0.4, \\
Q(s_{21}, \text{up}) &\leftarrow 0.5 + 0.1 \times 0.4 = 0.54.
\end{align*}
%
The estimate moved from $0.5$ toward the target of $0.9$ -- not
all the way, but a step in the right direction.
\Cref{fig:td-update} shows this update spatially: the value
estimate at $s_{22}$ bootstraps back to refine the estimate at
$s_{21}$. After many such updates across many transitions, the
estimates converge to $q_*$.

\input{figures/td-update}

\smallskip
The connection to the Bellman optimality equation
(\cref{eq:bellman-optimality-q}) is direct: the TD target
$R_{t+1} + \gamma \max_a Q(S_{t+1}, a)$ is a single-sample
estimate of the right-hand side of that equation. Each update
nudges the Q-values toward satisfying the Bellman optimality
equation, using real experience instead of known dynamics.

\smallskip
There is a tension, however, between learning and acting. If the
agent always picks the action with the highest Q-value -- the
\emph{greedy} action -- it will exploit what it currently knows
but may never discover that a different action is better. In the
grid-world, an agent that found a path to the $+1$ reward early
on might follow it forever, never exploring the route that leads
to a higher reward elsewhere. \cref{fig:exploration} shows this
tension in miniature: a greedy agent stays with the known $+1$
and never discovers the $+3$.

\input{figures/exploration}

\smallskip
This is the \emph{exploration--exploitation} trade-off: the agent
must sometimes try actions that look suboptimal in order to gather
information that could improve its long-term behavior. The
simplest solution is the $\epsilon$\emph{-greedy} strategy: with
probability $1 - \epsilon$ the agent takes the greedy action, and
with probability~$\epsilon$ it picks a random action instead.
Early in training, $\epsilon$ is large so the agent explores
broadly; over time, $\epsilon$ is decayed toward a small value so
the agent increasingly exploits its learned Q-values.

\smallskip
A notable property of Q-learning is that it is \emph{off-policy}:
the update rule always uses $\max_a Q(S_{t+1}, a)$ -- the value
of the best action -- regardless of which action the agent
actually takes next. The agent explores with $\epsilon$-greedy,
but the update targets the greedy (optimal) policy. This
separation is what makes exploration safe: the agent can try
random actions to discover new information without corrupting its
estimate of $q_*$.

\smallskip
Under standard conditions -- all state--action pairs continue to
be visited, and the learning rate is decreased
appropriately -- Q-learning converges to $q_*$ with probability
one~\citep{watkins1992q}. The $\epsilon$-greedy policy ensures
the visitation condition as long as $\epsilon > 0$.

\smallskip
The algorithm as described maintains a table with one entry for
every state--action pair. For the grid-world, this table has
$11 \times 4 = 44$ entries -- perfectly manageable. But for
problems with large state spaces, the table becomes impractical.
If the state is a raw image, the number of possible states grows
astronomically, and the agent would need to visit each one
repeatedly to learn its Q-values. Worse, learning that one state
is valuable tells the agent nothing about similar states it has
not yet seen -- there is no generalization.

\smallskip
The next section addresses this limitation by replacing the table
with a parameterized function that can generalize across states,
allowing Q-learning to scale to problems where explicit
enumeration is impossible.

\subsection{Function Approximation}
\label{subsec:func-approx}

Q-learning as described in the previous section maintains a table
with one entry for every state--action pair. When the state is a
raw image -- say, $84 \times 84$ grayscale pixels -- the number of
possible states is astronomical. A table cannot fit in memory, the
agent cannot visit every entry, and learning that one state is
valuable tells it nothing about similar states it has not yet seen.
Tabular Q-learning is not merely slow in such settings; it is
impossible.

\smallskip
The solution is to replace the table with a parameterized function.
Instead of storing a separate value for each state--action pair, we
define a function $Q(s, a;\, \mathbf{w})$ that takes a state and
action as input and produces a Q-value estimate as output, where
$\mathbf{w}$ is a vector of adjustable parameters (weights).

\smallskip
The key advantage is generalization. Because the function maps
states to Q-values through shared parameters, updating
$\mathbf{w}$ based on one state automatically changes the estimates
for similar states. In the grid-world analogy, this is like learning
the rule ``cells closer to the goal are worth more'' rather than
memorizing the value of each cell individually -- the rule applies
even to cells the agent has never visited.
\Cref{fig:table-vs-network} contrasts the two approaches.

\input{figures/table-vs-network}

\smallskip
To train the function, we need a loss to minimize. In supervised
learning, the loss compares predictions to known labels. Here, the
``labels'' are the TD targets from Q-learning: after each transition
$(S_t, A_t, R_{t+1}, S_{t+1})$, the target is
$R_{t+1} + \gamma \max_a Q(S_{t+1}, a;\, \mathbf{w})$ -- the same
quantity that appeared in the tabular update rule
(\cref{eq:q-learning}). The loss function is the squared difference
between the prediction and this target:
%
\begin{equation}
\label{eq:fa-loss}
L(\mathbf{w}) = \mathbb{E}\Big[\big(
  R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a;\, \mathbf{w})
  - Q(S_t, A_t;\, \mathbf{w})\big)^2\Big].
\end{equation}
%
Each transition provides one training example:
$Q(S_t, A_t;\, \mathbf{w})$ is the prediction, the TD target is
the label, and the squared TD error measures how far off the
prediction was.

\smallskip
Minimizing this loss follows the standard recipe from supervised
learning: compute the gradient of $L$ with respect to $\mathbf{w}$
and take a step in the direction that reduces the error. In
practice, we use stochastic gradient descent -- updating
$\mathbf{w}$ after each transition rather than averaging over many
transitions first:
%
\begin{equation}
\label{eq:sgd-update}
\mathbf{w} \;\leftarrow\; \mathbf{w}
  + \alpha \Big[R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a;\, \mathbf{w})
  - Q(S_t, A_t;\, \mathbf{w})\Big]\,
  \nabla_{\mathbf{w}}\, Q(S_t, A_t;\, \mathbf{w}).
\end{equation}
%
This is the function approximation counterpart of the tabular
Q-learning update. The TD error -- the bracketed term -- plays the
same role as before, but instead of directly adjusting one table
entry, it is scaled by the gradient
$\nabla_{\mathbf{w}}\, Q(S_t, A_t;\, \mathbf{w})$, which
distributes the correction across all parameters.

\smallskip
There is a subtlety: the TD target itself depends on $\mathbf{w}$,
because $Q(S_{t+1}, a;\, \mathbf{w})$ uses the same parameters we
are updating. A true gradient would differentiate through the target
as well, but in practice we treat the target as a fixed constant
when computing the gradient. This is called a \emph{semi-gradient}
method~\citep{sutton2018reinforcement}. Semi-gradient updates are
simpler and more stable than full-gradient alternatives, though they
weaken the theoretical convergence guarantees compared to supervised
learning. \Cref{fig:semi-gradient} illustrates this asymmetry: the
same parameters appear on both sides of the loss, but the gradient
update treats the target as if it were a fixed label.

\input{figures/semi-gradient}

\smallskip
For high-dimensional inputs such as images, neural networks are the
natural choice of function approximator. A neural network can learn
hierarchical features -- from edges and textures to shapes and
spatial arrangements -- directly from raw pixel values, without
requiring hand-crafted feature engineering.

\smallskip
In the architecture most relevant to what follows, the network takes
a raw image as input, processes it through convolutional layers that
extract spatial features, and feeds the result into fully connected
layers that output one Q-value per action. The entire network is
trained end-to-end by minimizing the loss in \cref{eq:fa-loss}, with
the gradient flowing from the output Q-values back through every
layer -- including the convolutional feature extractor. This means
the network does not just approximate Q-values; it also learns its
own representation of the state.

\smallskip
Function approximation makes Q-learning applicable to problems that
tabular methods cannot touch: the agent can process raw images,
generalize across similar states, and scale to environments with
vast state spaces. But these gains come at a cost. Tabular
Q-learning converges to $q_*$ under standard conditions; with
nonlinear function approximation, no such guarantee exists in
general.

\smallskip
The instability arises from the combination of three elements,
known as the \emph{deadly triad}~\citep{sutton2018reinforcement}.
First, \emph{bootstrapping}: the TD target depends on the current
estimates, so errors in the Q-values feed back into the targets
used to train them. Second, \emph{off-policy learning}: the $\max$
operator in the target means the agent updates toward a different
policy (the greedy one) than the one it uses to collect data,
which can amplify estimation errors. Third, \emph{function
approximation} itself: because parameters are shared across states,
a single update changes Q-values for many states simultaneously,
and these coupled changes can reinforce each other and cause the
estimates to diverge.

\smallskip
The next section introduces the deep Q-network (DQN), which
overcomes this instability through two stabilization
techniques -- experience replay and target networks -- producing
the first reinforcement learning agent to learn directly from raw
pixels.

\section{Deep Q-Networks}
\label{sec:dqn}

The deep Q-network (DQN) demonstrated that Q-learning with
function approximation could work at scale, learning to play
dozens of video games directly from raw pixels at
human-competitive levels~\citep{mnih2015human}. The algorithm
is Q-learning with a convolutional neural network as the
function approximator, plus two techniques that tame the
instability flagged in the previous section: \emph{experience
replay}, which breaks the correlations between consecutive
training samples, and a \emph{target network}, which stabilizes
the learning target. This section describes each component in
turn.

\subsection{Convolutional Encoder}
\label{subsec:conv-encoder}

The previous section described function approximation in general
terms: replace the Q-table with a neural network that processes
observations and outputs Q-values. This section makes that
concrete -- what does the network look like, and what does it
actually see?

\smallskip
The raw output of a video game environment is a sequence of color
frames -- $210 \times 160$ pixels with three color channels, rendered
at 60 frames per second. Feeding these directly into a neural network
would be wasteful: neighboring frames are nearly identical, color
carries little task-relevant information, and the resolution is
higher than needed. A four-step preprocessing pipeline reduces each
frame to a compact representation before the network sees it.

\smallskip
First, the agent takes the pixelwise maximum of two consecutive raw
frames. Some game environments render certain sprites on alternating
frames -- an object may appear in one frame and vanish in the next
due to hardware limitations in how sprites are drawn. The max
operation ensures that all objects are visible in every processed
frame.

\smallskip
Second, the frame is converted to grayscale by extracting the
luminance channel. For most games, the spatial layout of objects
matters far more than their color, and reducing from three channels
to one cuts the input size by two-thirds.
\Cref{fig:grayscale} illustrates the transformation.

\input{figures/grayscale-conversion}

\smallskip
Third, the frame is resized to $84 \times 84$ pixels, balancing
spatial resolution against computational cost. This size preserves
enough spatial detail for game play while keeping the input
manageable. \Cref{fig:resize} shows the spatial reduction.

\input{figures/resize}

\smallskip
Fourth, the four most recent preprocessed frames are stacked along
the channel dimension, producing an $84 \times 84 \times 4$ input
tensor. A single frame is a static image -- it shows positions but
not velocities. Stacking four frames gives the network access to
motion information: the direction and speed of a moving object can
be inferred from its positions across consecutive frames, without
requiring recurrence or explicit motion estimation
(\cref{fig:frame-stacking}).
\Cref{fig:preprocessing} summarizes the four steps.

\input{figures/frame-stacking}

\input{figures/preprocessing}

\smallskip
This preprocessing pipeline has become standard in deep reinforcement
learning for visual environments -- virtually every subsequent
method, including those discussed in later sections, uses the same
four steps~\citep{mnih2015human}.

\smallskip
The preprocessed tensor is the input to the \emph{Nature CNN}, a
convolutional neural network introduced by~\citet{mnih2015human}.
The architecture has three convolutional layers followed by a fully
connected layer and a linear output layer. Each convolutional layer
applies a set of learned filters that slide across the input,
detecting spatial patterns at progressively higher levels of
abstraction -- edges and textures in the first layer, shapes and
object parts in deeper layers. \Cref{fig:conv-operation} illustrates the basic convolution
operation: a kernel of learned weights slides across the input,
computing a weighted sum at each position to produce one output value.

\input{figures/conv-operations}

\smallskip
The three convolutional layers use 32, 64, and 64 filters with
kernel sizes of $8 \times 8$, $4 \times 4$, and $3 \times 3$ and
strides of 4, 2, and 1, respectively. The large kernel and stride in
the first layer rapidly reduce the spatial dimensions from
$84 \times 84$ to $20 \times 20$, while the later layers refine
features at finer spatial scales. \Cref{fig:stride-effect}
illustrates the trade-off: a larger stride means the kernel visits
fewer positions and the output shrinks accordingly.

\input{figures/stride-comparison}

\smallskip After the final convolutional
layer, the output is flattened and passed through a fully connected
layer with 512 units. Each hidden layer uses a ReLU activation
function. The output layer produces one Q-value for every available
action -- typically 18 for standard game environments.
\Cref{fig:nature-cnn} shows the architecture visually, and
\cref{tab:nature-cnn-layers} lists the exact specifications.

\input{figures/nature-cnn}
\input{tables/nature-cnn-layers}

\smallskip
This all-actions output design is a deliberate architectural choice.
Rather than taking a state--action pair as input and outputting a
single Q-value, the network takes only the state and produces all
Q-values in one forward pass. Computing
$\max_a Q(s, a;\, \mathbf{w})$ -- needed for both the greedy policy
and the TD target -- therefore requires a single pass through the
network, not $|\mathcal{A}|$ separate passes.

\smallskip
The convolutional layers are more than a fixed preprocessing step --
they are a learned feature extractor. The 512-dimensional activation
of the fully connected layer serves as a compressed representation
of the game state, capturing whatever spatial structure is relevant
for predicting Q-values. This representation is shaped entirely by
the TD error from \cref{eq:fa-loss}: the Q-learning loss is the only
training signal that drives the encoder. Whether this signal alone
produces representations that capture the full structure of the
environment, or whether additional self-supervised objectives can
improve them, is a central question in data-efficient reinforcement
learning.

\subsection{Experience Replay}
\label{subsec:experience-replay-bg}

The previous section described the architecture that maps raw frames
to Q-values. But training this network with standard Q-learning
updates -- applying the SGD rule from \cref{eq:sgd-update} to each
transition as it arrives -- is unstable. The problem lies not in the
network itself, but in the data it trains on.

\smallskip
In supervised learning, SGD convergence relies on the training data
being independent and identically distributed (i.i.d.): each example
in a minibatch is drawn independently from the same underlying
distribution. In reinforcement learning, this assumption is violated
in two ways.

\smallskip
First, consecutive transitions are highly correlated. If the agent
is in a particular state at time~$t$, the state at time~$t+1$ is
almost identical -- a ball has moved a few pixels, or a character has
shifted slightly. Training on a sequence of near-identical inputs
pushes the network's parameters in the same direction repeatedly,
overfitting to the agent's recent local experience rather than
learning general value estimates.

\smallskip
Second, the data distribution itself changes as the agent learns.
Each gradient step updates the network, which changes the policy,
which changes which states the agent visits, which changes the
distribution of training data. The agent is training on a shifting
distribution of its own creation -- a feedback loop that can cause
the network's predictions to diverge.

\smallskip
\emph{Experience replay}~\citep{lin1992self,mnih2013playing} breaks
this loop with a simple mechanism: instead of training on each
transition once as it arrives, the agent stores every transition
$(s, a, r, s')$ in a large buffer and trains on random minibatches
sampled uniformly from the buffer. The buffer is a fixed-size
first-in-first-out queue -- typically holding one million transitions.
When full, the oldest transitions are discarded to make room for
new ones. \Cref{fig:replay-buffer} illustrates the mechanism:
transitions arrive sequentially from the agent, but the training
minibatch is assembled by sampling randomly across the buffer.

\input{figures/replay-buffer}

\smallskip
Random sampling from the buffer serves three purposes. First, it
\emph{decorrelates} the training data: a minibatch drawn from a
million-transition buffer contains experiences from many different
episodes and time steps, approximating the i.i.d.\ condition that
SGD requires.

\smallskip
Second, it improves \emph{data efficiency}: each transition can
appear in many minibatches over the course of training, rather than
being used for a single gradient step and discarded. This is
especially important when environment interactions are expensive --
extracting more learning from each transition means the agent needs
fewer total interactions to improve.

\smallskip
Third, it provides \emph{smoothing}: because the buffer accumulates
transitions from many past versions of the policy, the training
distribution changes gradually rather than shifting abruptly with
each policy update. This prevents the network from overfitting to
its most recent behavior and reduces oscillations in the learned
values.

\smallskip
Experience replay connects directly to the function approximation
loss in \cref{eq:fa-loss}. That equation defined the expected
squared TD error, with the expectation taken over transitions --
but did not specify where those transitions come from. Experience
replay provides the answer: the expectation is approximated by
sampling minibatches from the buffer, and each sampled transition
contributes one term to the loss.

\subsection{Target Networks}
\label{subsec:target-networks-bg}

Experience replay solves the data correlation problem, but a second
instability remains. Recall from \cref{subsec:func-approx} that the
TD target in the semi-gradient update depends on the same parameters
$\mathbf{w}$ that are being updated: the target
$R_{t+1} + \gamma \max_a Q(S_{t+1}, a;\, \mathbf{w})$ shifts every
time the network takes a gradient step. The agent is optimizing
toward a goal that moves with every update.

\smallskip
In supervised learning, labels are fixed -- the loss landscape does
not change during training, and SGD can make steady progress toward
a minimum. In Q-learning with function approximation, the
``labels'' are the TD targets, and they depend on the current
parameters. Every step that improves the prediction also shifts the
target, creating a feedback loop: the network's estimates and its
targets chase each other, and training can oscillate or diverge
rather than converge. This is the bootstrapping element of the
deadly triad -- the target depends on the very estimate being
refined.

\smallskip
The solution is to decouple the prediction from the target by
maintaining two copies of the network. The \emph{online network},
with parameters~$\boldsymbol{\theta}$, is the network being
trained -- it receives gradient updates at every step. The
\emph{target network}, with parameters~$\boldsymbol{\theta}^-$, is
a frozen copy used only to compute the TD target. The target
becomes:
%
\[
R_{t+1} + \gamma \max_{a} Q(S_{t+1}, a;\,
  \boldsymbol{\theta}^-),
\]
%
where $\boldsymbol{\theta}^-$ does not change during training.
Because the target is fixed, the loss landscape is stationary and
SGD can make progress toward a stable objective -- the same
conditions that make supervised learning work.

\smallskip
Periodically -- every $C$ gradient steps -- the target network is
replaced with a copy of the online network:
$\boldsymbol{\theta}^- \leftarrow \boldsymbol{\theta}$. In the
original DQN, $C = 10{,}000$. Between these updates, the target
network is completely frozen: no gradients flow through it, and its
parameters do not change. \Cref{fig:target-network} illustrates this
asymmetry: gradients update only the online network, while the target
network provides a stable reference.

\input{figures/target-network}

\smallskip
With both stabilization techniques in place, the complete DQN loss
combines experience replay and the target network:
%
\begin{equation}
\label{eq:dqn-loss}
L(\boldsymbol{\theta}) = \mathbb{E}_{(s,a,r,s') \sim \mathcal{D}}
  \Big[\big(r + \gamma \max_{a'} Q(s', a';\,
  \boldsymbol{\theta}^-) - Q(s, a;\, \boldsymbol{\theta})
  \big)^2\Big].
\end{equation}
%
This is the function approximation loss from \cref{eq:fa-loss} with
two concrete changes. The expectation is over minibatches sampled
uniformly from the replay buffer~$\mathcal{D}$, addressing the data
correlation problem. And the target uses frozen
parameters~$\boldsymbol{\theta}^-$ instead of the current
parameters~$\boldsymbol{\theta}$, stabilizing the optimization
target. These two changes -- replay and frozen targets -- are DQN's
core algorithmic contribution, and they are what make Q-learning
with a deep neural network stable enough to
learn~\citep{mnih2015human}.

\smallskip
The update interval $C$ controls a trade-off between stability and
staleness. A longer interval gives more stability -- the target
moves less frequently, and SGD has more time to converge toward it.
But it also makes the target more stale, since
$\boldsymbol{\theta}^-$ falls further behind
$\boldsymbol{\theta}$ as training progresses. Too long, and the
agent trains against an outdated target that no longer reflects its
current knowledge; too short, and the target moves so often that the
stability benefit is lost.

\smallskip
Together, experience replay and target networks address two of the
three instabilities in the deadly triad: correlated training data
and a shifting optimization target. They do not eliminate the third
element -- the use of a nonlinear function approximator -- but they
reduce the practical instability enough for learning to succeed.
The next section introduces Rainbow, which extends DQN with six
additional improvements that further boost performance.

\subsection{Rainbow DQN}
\label{subsec:rainbow-dqn}

Between 2015 and 2018, several independent lines of research each
identified a specific weakness in DQN and proposed a targeted fix.
\citet{hessel2018rainbow} showed that combining six of these
improvements into a single agent -- \emph{Rainbow} -- produces
performance far exceeding any individual extension. The gains are
largely complementary, not redundant: each improvement addresses a
different aspect of learning, and their benefits compound when used
together. \Cref{tab:rainbow-extensions} summarizes the six
extensions; the paragraphs below describe each one.

\input{tables/rainbow-extensions}

\smallskip
\textbf{Double DQN.}\quad The $\max$ operator in the Q-learning
target (\cref{eq:dqn-loss}) causes a systematic upward bias: when
Q-value estimates are noisy, taking the maximum over noisy values
tends to select overestimates. Double
DQN~\citep{van2016deep} fixes this by decoupling action
selection from action evaluation. The online network selects the
best action, but the target network evaluates it -- reducing the
overestimation because two independent estimates are unlikely to
both be biased in the same direction for the same action.

\smallskip
\textbf{Prioritized experience replay.}\quad Standard experience
replay samples transitions uniformly from the buffer, but not all
transitions are equally informative. Prioritized
replay~\citep{schaul2016prioritized} samples transitions with
probability proportional to their TD error -- the magnitude of the
difference between the predicted and target Q-values. Transitions
where the agent's prediction was furthest off are replayed more
often, focusing learning on the most surprising or informative
experiences.

\smallskip
\textbf{Dueling networks.}\quad Rather than estimating $Q(s, a)$
directly, the dueling architecture~\citep{wang2016dueling}
decomposes the Q-value into a state-value $V(s)$ and an advantage
$A(s, a)$, combining them as
$Q(s, a) = V(s) + A(s, a) - \overline{A}(s)$. Both streams share
the convolutional encoder but have separate fully connected heads.
This separation lets the network learn ``how good is this state''
independently from ``how much better is this action than average,''
which is beneficial in states where the choice of action matters
little. \Cref{fig:dueling-architecture} contrasts the two
architectures.

\input{figures/dueling-architecture}

\smallskip
\textbf{Multi-step returns.}\quad Standard DQN bootstraps from the
immediate next state -- a one-step target. Multi-step
returns~\citep{sutton2018reinforcement} instead use the actual
rewards from the next $n$ steps before bootstrapping, computing the
target as
$\sum_{k=0}^{n-1} \gamma^k R_{t+k+1}
  + \gamma^n \max_a Q(S_{t+n}, a)$.
This propagates reward information faster and shifts the
bias--variance trade-off: more real rewards and less bootstrapping
reduces bias at the cost of higher variance. Rainbow uses $n = 3$.

\smallskip
\textbf{Distributional RL.}\quad Instead of estimating the expected
return -- a single number -- distributional
RL~\citep{bellemare2017distributional} models the full distribution
of possible returns. The C51 variant used in Rainbow represents this
distribution as a probability mass function over 51 evenly spaced
values. The network outputs probabilities for each value, and the
loss minimizes the KL divergence between the predicted and target
distributions. This richer learning signal can improve feature
learning even though the policy still acts on the expected value.
\Cref{fig:distributional-output} illustrates the difference in
output structure.

\input{figures/distributional-output}

\smallskip
\textbf{Noisy networks.}\quad DQN uses $\varepsilon$-greedy
exploration, which selects random actions with fixed
probability~$\varepsilon$ regardless of the agent's uncertainty.
Noisy networks~\citep{fortunato2018noisy} replace this with learned
parametric noise injected into the network's weights. The network
learns to explore where it is uncertain and exploit where it is
confident, rather than exploring uniformly at random. With noisy
networks, $\varepsilon$ is set to zero -- action selection is fully
greedy with respect to the (noisy) Q-values.

\smallskip
Taken together, these six extensions modify nearly every component
of DQN: the loss function changes from squared error to
distributional KL divergence, the replay strategy from uniform to
prioritized sampling, the target computation from single-step to
multi-step with double Q-learning, the exploration strategy from
$\varepsilon$-greedy to noisy networks, and the network head from
a single stream to the dueling architecture. The only component
left unchanged is the convolutional encoder -- the Nature CNN
backbone from \cref{subsec:conv-encoder}.
\Cref{fig:rainbow-components} maps each extension to the component
it modifies, highlighting the encoder as the sole unchanged piece.

\input{figures/rainbow-components}

\smallskip
\citet{hessel2018rainbow} performed ablation experiments, removing
one component at a time to measure each extension's contribution.
The most critical components were prioritized replay and multi-step
returns; the least impactful in isolation were double Q-learning and
dueling networks. However, this ranking was established at 200
million frames of training. Whether the same ranking holds in the
data-efficient regime -- where the agent has far fewer interactions
to learn from -- is less clear, as the relative importance of each
component may shift when data is scarce.

\smallskip
That the convolutional encoder remains unchanged across all of these
improvements is significant. Methods such as SPR, introduced in
\cref{subsec:spr}, add self-supervised objectives that operate
directly on the encoder's learned representations. When SPR was
shown to improve Rainbow's performance, it was unclear whether the
gains came from SPR's representation learning or from Rainbow's many
enhancements providing a richer learning substrate. Disentangling
these factors -- testing whether SPR's benefits transfer to a
simpler base agent -- is one of the questions that motivates the
experiments in this work.

\section{The Arcade Learning Environment}
\label{sec:ale}

The previous section introduced two agents -- vanilla DQN and
Rainbow -- that differ in algorithmic sophistication but share the
same convolutional encoder and the same type of input: preprocessed
Atari game frames. This section introduces the benchmark on which
both agents are evaluated. The Arcade Learning
Environment~\citep{bellemare2013arcade} provides a standardized
interface to dozens of Atari 2600 games, making it possible to
evaluate a single agent architecture across a diverse set of tasks
without game-specific engineering. We describe the standard
evaluation protocol that established DQN's original results, the
data-efficient 100K variant used in our experiments, and the
metrics used to compare agents across games.

\subsection{The Standard Protocol}
\label{subsec:standard-protocol}

\citet{bellemare2013arcade} introduced the Arcade Learning
Environment (ALE): a software framework that wraps Atari 2600 game
ROMs into a standardized reinforcement learning interface. For
each game, the environment exposes $210 \times 160$ RGB frames
rendered at 60 frames per second, a discrete action set of up to
18 actions, a scalar reward signal derived from the game score, and
an episode termination signal. The key property is uniformity:
every game uses the same interface, so a single agent architecture
can be evaluated across the full suite without modification.

\smallskip
The games span a wide range of challenges -- spatial reasoning and
reaction time in Breakout, navigation and long-horizon planning in
Montezuma's Revenge, precise timing in Freeway, and multi-object
tracking in Space Invaders. This diversity is what made ALE the
standard benchmark for deep reinforcement learning: strong
performance across many games requires general-purpose
representations, not game-specific heuristics.

\smallskip
In the standard protocol, the agent does not act on every raw
frame. Instead, it selects an action every $k$-th frame, and the
chosen action is repeated for the intermediate frames.
\citet{mnih2015human} used $k = 4$, which became the default. This
\emph{frame skip} (or \emph{action repeat}) means that one agent
decision corresponds to four raw frames. The distinction between
raw frames and agent decisions matters when comparing training
budgets: 200 million raw frames is 50 million environment steps.
Frame skip serves two purposes: consecutive frames are nearly
identical, so acting on every one is unnecessary, and skipping
frames effectively speeds up the game from the agent's perspective.

\smallskip
The preprocessing pipeline that converts raw frames into the
$84 \times 84 \times 4$ input tensor was described in
\cref{subsec:conv-encoder}. This pipeline -- max-pooling over
consecutive frames, grayscale conversion, resizing, and stacking
four frames -- is universal across methods. DQN, Rainbow, DrQ,
and SPR all receive the same preprocessed observations.

\smallskip
\citet{mnih2015human} trained DQN for 200 million frames per game,
roughly 38 days of game time at 60 frames per second. Each game
was trained independently with a single set of hyperparameters
shared across the entire suite. Evaluation followed a fixed
procedure: insert a random number (up to 30) of no-op actions at
the start of each episode to randomize the initial state, then let
the agent play for up to five minutes. Performance was averaged
over 30 evaluation episodes. Under this protocol, DQN achieved
human-competitive scores on roughly 60\% of 49 games -- the first
demonstration that a single architecture could learn diverse
visuomotor tasks from raw pixels.

\smallskip
\citet{machado2018revisiting} later identified weaknesses in the
original protocol and proposed two revisions that are now standard.
The first is \emph{sticky actions}: with probability 0.25, the
environment repeats the agent's previous action instead of
executing its current choice. Many Atari games are deterministic --
the same sequence of actions always produces the same outcome. An
agent trained on a deterministic game can exploit this by
memorizing fixed action sequences that achieve high scores without
learning general strategies. Sticky actions inject stochasticity
into the transition dynamics, forcing the agent to cope with
variability and preventing such memorization.

\smallskip
The second revision concerns episode boundaries. The original
protocol treated each life loss as an episode termination,
resetting the agent's trajectory. While this helped in some games
by making early deaths less costly to explore from, it changed the
effective task: the agent optimized for one-life performance rather
than full-game performance. \citet{machado2018revisiting}
recommended using only the game-over signal for episode boundaries,
preserving the original game structure. Both revisions -- sticky
actions and game-over termination -- are adopted in the 100K
protocol and in our experiments.

\smallskip
This protocol -- 200 million frames per game -- gave DQN enough
experience to learn effective policies for most games. But from a
practical standpoint, 200 million frames is an enormous amount of
interaction. The next section introduces a protocol that asks
whether agents can learn with far less.

\subsection{The 100K Protocol}
\label{subsec:100k-protocol}

\citet{kaiser2020model} proposed the Atari-100K benchmark: agents
are limited to 100{,}000 environment steps per game. With frame
skip $k = 4$, this corresponds to 400{,}000 raw frames -- roughly
six and a half minutes of game time at 60 frames per second.
Compared to the standard 200 million frames, the agent sees 0.2\%
of the data that DQN was originally trained on: a reduction of
500$\times$.

\smallskip
This constraint is not merely a smaller version of the standard
protocol. It creates a qualitatively different learning regime. In
the 200-million-frame setting, an agent can afford to explore
randomly for millions of steps, encounter diverse game situations
through sheer volume of experience, and gradually refine its
representations. At 100{,}000 steps, none of this is possible. The
agent must learn useful representations from its first experiences
and extract as much information as possible from every transition.
Vanilla DQN, which was designed for the high-data regime, barely
learns anything on most games under this constraint.

\smallskip
The benchmark has practical motivation beyond evaluating
algorithms. In real-world domains -- robotics, autonomous driving,
clinical decision support -- collecting millions of interactions is
expensive, slow, or dangerous. An agent that learns from 100{,}000
steps is closer to practical deployment than one requiring 200
million. The 100K benchmark distills this challenge into a
controlled setting where methods can be compared systematically.

\smallskip
The benchmark selects 26 games from the full 57-game ALE suite,
chosen to span a range of challenges: reward density, visual
complexity, planning horizon, and interaction patterns. Some games
offer frequent rewards that guide learning (Breakout, Pong), while
others require extended exploration before any reward is
encountered (Montezuma's Revenge, Private Eye). This diversity
ensures that strong aggregate performance requires general-purpose
data efficiency, not a strategy that works for only one type of
game.

\smallskip
After training for 100{,}000 steps, the agent is evaluated over
100 episodes using a near-greedy policy ($\varepsilon = 0.001$).
The evaluation score is the average undiscounted return across
these episodes. Sticky actions ($p = 0.25$) apply during both
training and evaluation, following the revised protocol of
\citet{machado2018revisiting}. The original benchmark paper
\citep{kaiser2020model} did not use sticky actions, but subsequent
work -- including SPR~\citep{schwarzer2021dataefficient} and the
Dopamine framework -- adopted them. Our experiments use sticky
actions consistently.

\smallskip
Hyperparameters tuned for 200 million frames are poorly suited to
the 100K regime. Key adjustments include a higher replay ratio --
performing more gradient updates per environment step to extract
more learning from each transition -- a shorter warmup period
before training begins, and faster $\varepsilon$ decay to shift
from exploration to exploitation earlier given the tiny budget.
\citet{kielak2020recent} demonstrated this concretely: the
model-free Rainbow baseline reported in the original benchmark
paper used default Dopamine hyperparameters without
game-specific tuning, making model-free methods appear weaker than
they are. A properly tuned Rainbow (``OTRainbow'') substantially
outperformed the original baseline without any algorithmic changes.
Published 100K results are sensitive to hyperparameter choices, not
just algorithmic differences -- a point that matters when comparing
across papers.

\smallskip
Comparing agents under the 100K constraint requires metrics that
aggregate performance across games with very different score
ranges. A score of 500 in one game might represent near-perfect
play, while the same score in another game is negligible. The next
section introduces the normalization and aggregation methods that
make these comparisons meaningful.

\subsection{Metrics}
\label{subsec:metrics}

Raw scores are not comparable across games. Pong scores range from
$-21$ to $+21$; Crazy Climber scores reach hundreds of thousands.
Averaging raw scores would let a single high-scoring game dominate
the aggregate, hiding poor performance elsewhere. Meaningful
comparison requires a normalization that puts all games on a common
scale.

\smallskip
The standard normalization is the \emph{human-normalized score}
(HNS):
\begin{equation}
\label{eq:hns}
\text{HNS} = \frac{\text{agent score} - \text{random score}}
                   {\text{human score} - \text{random score}}.
\end{equation}
An HNS of 0 means the agent performs at the level of a random
policy; an HNS of 1 means it matches human performance; values
above 1 are superhuman. In concrete terms, HNS measures what
fraction of the gap between random play and human play the agent
closes. The human and random baselines come from
\citet{mnih2015human}: human scores were recorded by a
professional games tester playing each game for two hours, and
random scores were measured from a uniform random policy.

\smallskip
Once scores are normalized, they must be aggregated across games
into a single number for comparison. The two obvious choices --
mean and median -- each have problems.
The mean is sensitive to outliers: if an agent achieves an HNS of
10 on one game (far above human level) but 0.1 on the rest, the
mean looks strong despite broadly weak performance. The median is
robust to outliers but statistically inefficient -- it uses only
the middle value and discards most of the data.
\citet{agarwal2021deep} demonstrated the practical consequences:
with only five runs per game (common in the literature), the median
of the same algorithm can vary by over 30\% depending on which
random seeds are included.

\smallskip
\citet{agarwal2021deep} recommended the \emph{interquartile mean}
(IQM) as the primary aggregate metric. IQM is a 25\% trimmed mean:
pool the normalized scores from all game-run combinations, discard
the bottom 25\% and top 25\%, and take the mean of the remaining
50\%. This makes IQM more robust than the mean (outliers are
trimmed) and more statistically efficient than the median (it uses
half the data rather than a single point), producing tighter
confidence intervals for the same number of runs.

\smallskip
Point estimates of any aggregate metric are unreliable when the
number of runs is small. \citet{agarwal2021deep} proposed reporting
\emph{stratified bootstrap confidence intervals}: for each game,
resample the available runs with replacement; compute the aggregate
metric on each bootstrap sample; report the 95\% interval. This
tells the reader how much the aggregate could shift with different
random seeds. A difference between two agents is meaningful only if
their confidence intervals show clear separation.

\smallskip
A complementary tool is the \emph{performance profile}: a curve
showing the fraction of game-run combinations where the agent
exceeds a given normalized score threshold. Higher curves indicate
stronger algorithms, and a curve that lies strictly above another
indicates stochastic dominance across the full range of thresholds.
Performance profiles provide a richer picture than any single
aggregate metric because they reveal the full distribution of
performance -- not just a summary statistic.

\smallskip
These metrics -- human-normalized scores aggregated via IQM with
bootstrap confidence intervals -- provide a principled framework
for comparing agents across games with very different score
ranges. The next section introduces data augmentation, the first
of two techniques that improve data efficiency under the 100K
constraint.

\section{Data Augmentation for RL}
\label{sec:augmentation}

The previous section established the 100K constraint: agents must
learn from roughly six minutes of play. Under such a tight data
budget, the agent replays the same transitions repeatedly,
risking overfitting to specific pixel patterns rather than
learning general features. Data augmentation -- a standard
technique in supervised learning -- has recently been shown to be
surprisingly effective in reinforcement learning. By applying
small, semantics-preserving transformations to game frames during
training, augmentation expands the effective dataset without
additional environment interaction. This section describes the
specific augmentation used in data-efficient RL (random spatial
shifts) and explains why it works as a form of implicit
regularization.

\subsection{Random Shift Augmentation}
\label{subsec:random-shift}

In supervised learning, data augmentation applies label-preserving
transformations to training inputs: random crops, horizontal flips,
color jitter. The key requirement is that the transformation must
not change the correct answer -- a horizontally flipped cat is
still a cat. Augmentation is standard practice for training image
classifiers, especially when data is limited, because it
effectively multiplies the dataset by creating many versions of
each training example.

\smallskip
Transferring this idea to reinforcement learning is not
straightforward. In supervised learning, the label for an image is
fixed: a cat is always a cat regardless of how many times the image
is transformed. In RL, the ``label'' for an observation is the
Q-value target, which depends on the current policy and changes
every training step. But the observation itself is still an image.
If a small spatial transformation does not change the game
semantics -- the paddle is in nearly the same position, the ball
trajectory is the same, the optimal action has not changed -- then
the Q-values should also be unchanged. Spatial transforms that
produce small, semantics-preserving perturbations should therefore
be safe for RL, despite the non-stationary target problem.

\smallskip
The specific augmentation used in data-efficient RL is the
\emph{random shift}. Starting from the preprocessed
$84 \times 84$ grayscale frame, the operation pads the image by 4
pixels on each side using reflection padding, creating a
$92 \times 92$ image where the border mirrors the edge pixels. A
random $84 \times 84$ crop is then taken from this padded image,
with the crop position sampled uniformly at random. The result is
the original game frame shifted by up to $\pm 4$ pixels in both the
horizontal and vertical directions. The visual content is preserved;
only the spatial position has changed slightly.
\Cref{fig:random-shift} illustrates the three steps. This is the
operation used by DrQ~\citep{kostrikov2020image}; the equivalent
operation in RAD~\citep{laskin2020reinforcement} is called ``random
translate.''

\input{figures/random-shift}

\smallskip
Because the observation is a stack of four consecutive frames, the
same random shift -- the same horizontal and vertical offset -- is
applied to all four frames in the stack. This consistency matters:
the agent infers velocity and motion from pixel differences between
consecutive frames. If each frame were shifted independently, these
inter-frame differences would be corrupted, creating spurious
motion signals. Consistent augmentation across the stack preserves
temporal information while still providing spatial perturbation.

\smallskip
During training, each transition $(s, a, r, s')$ sampled from the
replay buffer is augmented before computing the loss. The current
state $s$ and next state $s'$ receive independent random shifts --
different crop positions -- because they are separate observations.
There is no need for the spatial offset of $s$ and $s'$ to match.
Both DrQ and RAD apply augmentation in this way.

\smallskip
\citet{laskin2020reinforcement} systematically tested ten
augmentation types on pixel-based RL: crop, translate, window,
grayscale, cutout, cutout-color, flip, rotate, random convolution,
and color jitter. Random crop and translate were the most effective
by a wide margin; most other augmentations were neutral or harmful.
The reason is that game semantics are approximately
translation-invariant at small scales -- a paddle shifted three
pixels is still a paddle in approximately the same place -- but
they are not invariant to other transforms. Horizontal flips
reverse left and right, changing the optimal action. Rotation
disrupts spatial layout (gravity goes sideways). Color jitter
alters object appearance in ways the encoder should remain
sensitive to. Random shift succeeds because it perturbs position
without breaking the game's spatial structure.

\smallskip
DrQ~\citep{kostrikov2020image} and
RAD~\citep{laskin2020reinforcement} are concurrent papers that
independently demonstrated image augmentation is sufficient for
competitive data-efficient RL. DrQ additionally proposed averaging
Q-targets across $K$ augmented versions of $s'$ and averaging
Q-values across $M$ augmented versions of $s$, but the simplest
setting -- $K = 1$, $M = 1$, a single random shift per observation
-- is already sufficient for Atari-100K. No complex averaging or
multiple forward passes are needed.

\subsection{Augmentation as Regularization}
\label{subsec:aug-regularization}

In the 100K regime, the replay buffer contains at most around
100{,}000 transitions -- this is the agent's entire dataset. With a
replay ratio of one (one gradient update per environment step),
each transition is sampled roughly once. Data-efficient settings
often use higher replay ratios, reusing each transition two to four
times. When the same transitions are replayed repeatedly, the
network risks memorizing specific pixel patterns rather than
learning general features. This is the same overfitting problem
that afflicts small-dataset classifiers in supervised learning, but
amplified by the shifting target distribution in RL.

\smallskip
Consider what overfitting looks like at the representation level.
The encoder might learn that a bright pixel at position $(42, 17)$
predicts high Q-values -- because in the limited training data,
that pixel happens to be bright when the ball is near the paddle.
This feature is fragile: it breaks with any positional variation.
A robust encoder would instead learn to represent the ball's
position relative to the paddle, a feature that generalizes across
spatial shifts. Without augmentation, the encoder has no pressure
to prefer the robust feature over the fragile one, because both
predict Q-values equally well on the exact training observations.

\smallskip
Random shift augmentation addresses this directly. Each time a
transition is replayed, the observation is augmented with a
different random spatial shift. Even though the buffer contains
only 100{,}000 transitions, the network never sees the exact same
input twice. This effectively expands the training distribution:
each transition generates a family of slightly different training
examples, all sharing the same Q-value target because the game
semantics are unchanged. The encoder is forced to learn features
that are invariant to small spatial shifts -- features that capture
``where the paddle is relative to the ball'' rather than ``what
pixel values appear at specific coordinates.'' This is
\emph{implicit regularization}: not an explicit penalty in the loss
function, but a constraint on what the encoder can learn, imposed
by the data distribution it trains on.

\smallskip
The same mechanism explains why augmentation helps in supervised
learning: random crops force a classifier to recognize objects
regardless of their position in the frame. In RL, the effect is
amplified because the dataset is orders of magnitude smaller --
100{,}000 transitions versus millions of labeled images -- and
because the training distribution shifts as the policy improves,
compounding the risk of overfitting to early, unrepresentative
data.

\smallskip
CURL~\citep{srinivas2020curl} combined contrastive learning with
random crop augmentation on data-efficient Rainbow, reporting
improved performance on Atari-100K. The contrastive loss requires
augmentation to generate positive pairs: two different random crops
of the same observation form a positive pair, while crops of
different observations form negative pairs. Because augmentation is
integral to the contrastive objective, the original results could
not distinguish whether the gains came from contrastive learning or
from the augmentation itself. DrQ and RAD answered this question:
augmentation alone, without any contrastive or self-supervised
loss, achieves similar or better results. The augmentation was
doing most of the work. This finding is important for experimental
design: augmentation and representation learning can be studied as
independent factors only when the representation learning method
does not depend on augmentation for its training signal.

\smallskip
Augmentation ensures the encoder sees diverse views of each
observation, encouraging spatial invariance and reducing
overfitting. But it does not change what the encoder is
\emph{trained to predict}. The Q-learning loss remains the sole
objective shaping the encoder's representations, and in the 100K
regime this signal is weak -- targets are bootstrapped from poor
estimates early in training, providing noisy guidance about what
features matter. The encoder may settle on representations that
satisfy the current Q-targets but fail to support further learning.
This is the gap that self-supervised representation learning
addresses: an auxiliary training objective that shapes the
encoder's representations to capture dynamics and structure beyond
what the Q-learning loss alone requires.

\section{Self-Supervised Representation Learning}
\label{sec:ssl}

The previous section showed that data augmentation reduces
overfitting by diversifying the encoder's inputs. But augmentation
does not change what the encoder is trained to predict -- the
Q-learning loss remains the sole objective shaping the encoder's
representations. In the 100K regime, this signal is weak: Q-targets
are bootstrapped from the agent's own poor estimates, and the
encoder may learn shallow features that satisfy current targets
without capturing the full structure of the environment.
Self-supervised representation learning addresses this gap by adding
an auxiliary training objective that shapes the encoder's
representations independently of the reward signal. This section
introduces the general idea, the key technical challenge
(representational collapse), and the specific method used in our
experiments: Self-Predictive Representations
(SPR)~\citep{schwarzer2021dataefficient}.

\subsection{Learning Without Labels}
\label{subsec:learning-without-labels}

Recall the DQN architecture from \cref{subsec:conv-encoder}: raw
frames pass through three convolutional layers and a fully
connected layer, producing a 512-dimensional vector. This vector
is the agent's internal representation of the game state --
everything the agent ``knows'' about the current observation
passes through this bottleneck. The Q-values for all actions are
computed from this vector, and the Q-learning loss
(\cref{eq:dqn-loss}) is the only training signal that shapes it.
The encoder learns to extract whatever features predict Q-values,
and nothing more.

\smallskip
In the 100K regime, this single training signal is problematic.
Q-targets are bootstrapped from the agent's own estimates, which
are poor early in training. The encoder receives weak, noisy
gradient information about what features to extract. With limited
data, it may find features that satisfy the current Q-targets
without capturing the full structure of the environment -- learning,
for instance, that dark pixels at the top of the screen correlate
with high Q-values (because the ball happens to be there in the
small set of stored transitions), rather than learning to track the
ball's position and velocity more generally. This is overfitting at
the representation level: the features work for the data seen so
far but do not support generalization to new states.

\smallskip
In the standard 200-million-frame regime, this is less of a
problem. With enough data, the Q-learning loss eventually provides
sufficient gradient to learn good representations -- the encoder
sees enough diverse states that fragile features are weeded out.
The 100K regime does not offer this luxury. Good representations
must emerge quickly or not at all.

\smallskip
Good representations make downstream learning easier. If the
encoder captures object positions, velocities, and spatial
relationships, then the Q-head -- a single linear layer -- can
more easily learn to map these features to accurate Q-values. Poor
representations force the Q-head to compensate for missing
information, which is harder and requires more data. The question
is how to shape the encoder's representations with an additional
training signal that does not depend on the quality of the
Q-targets.

\smallskip
Self-supervised learning creates training signals from the
structure of the data itself, without requiring external labels. The
general principle is to define a \emph{pretext task}: an auxiliary
objective that requires the encoder to capture useful information
about the input. The pretext task uses no external labels -- it
derives supervisory signals from the observations themselves. The
term ``self-supervised'' distinguishes this approach from supervised
learning (which needs external labels like class names or reward
signals) and from unsupervised learning (which typically refers to
clustering or density estimation).

\smallskip
In computer vision, self-supervised methods have proven remarkably
effective. Early approaches defined pretext tasks such as
predicting image rotations, solving jigsaw puzzles from shuffled
patches, or colorizing grayscale images. The insight behind these
tasks is that solving them requires understanding the image's
structure: a network that can predict how an image was rotated has
implicitly learned to recognize the objects and spatial layout
within it. Modern methods such as SimCLR~\citep{chen2020simple} and
BYOL~\citep{grill2020bootstrap} learn representations by comparing
augmented views of the same image, and their learned features rival
those of supervised pre-training on downstream tasks.

\smallskip
Reinforcement learning has something that static image datasets do
not: temporal structure. The agent observes sequences of states
connected by actions -- state $s_t$, action $a_t$, and next state
$s_{t+1}$ form a natural prediction problem. Given the current
state and the action taken, predict the next state. This is richer
than spatial-only pretext tasks because it captures \emph{dynamics}
-- how the game state evolves under different actions. An encoder
that can predict the next state must represent information about
object positions, velocities, and the effects of actions: exactly
the features that are useful for decision-making.

\smallskip
The prediction does not need to happen in pixel space. Predicting
future pixels is expensive and requires modeling irrelevant details
-- exact textures, background patterns, rendering artifacts.
Predicting future \emph{representations} -- the encoder's own
internal features -- focuses on the abstract structure that matters.
This is the key idea behind SPR (\cref{subsec:spr}): train the
encoder to predict its own future latent states, not future images.

\smallskip
The self-supervised loss is added alongside the Q-learning loss as
an auxiliary objective:
\begin{equation}
\label{eq:ssl-loss}
\mathcal{L}_{\text{total}}
  = \mathcal{L}_{\text{Q}}
  + \lambda\, \mathcal{L}_{\text{SSL}}.
\end{equation}
The encoder receives gradients from both losses. The Q-learning
loss tells the encoder what matters for decisions (action values);
the self-supervised loss tells it what matters for understanding
the environment's dynamics. The two signals are complementary: the
Q-loss provides task-relevant guidance (even if noisy), while the
self-supervised loss provides structurally rich guidance (even if
not directly about rewards). The weight $\lambda$ controls the
balance between the two objectives. In SPR, $\lambda = 2$, giving
the self-supervised signal twice the weight of the Q-learning loss.
\Cref{fig:dual-objective} illustrates this dual-objective
architecture: the shared encoder feeds both the Q-head and the
self-supervised head, and gradients from both losses flow back
through the encoder.

\input{figures/dual-objective}

\smallskip
Self-supervised methods fall into two broad families.
\emph{Contrastive methods} (SimCLR, MoCo,
CURL~\citep{srinivas2020curl}) learn by pulling together
representations of similar inputs (positive pairs) and pushing
apart representations of different inputs (negative pairs).
\emph{Non-contrastive} or \emph{predictive methods} (BYOL, SPR)
learn by predicting target representations without negatives,
using an exponential moving average (EMA) target encoder to provide
stable prediction targets. SPR belongs to the second family. The
distinction matters for RL because non-contrastive methods are
simpler (no negative sampling), more robust to small batch sizes
(RL typically uses batches of 32, far smaller than the thousands
used in vision), and less dependent on augmentation design.

\smallskip
But prediction-based objectives have a fundamental obstacle: if the
encoder maps every input to the same constant vector, the
prediction task is solved perfectly -- with a useless
representation. The next section explains this collapse problem and
how modern methods overcome it.

\subsection{The Collapse Problem}
\label{subsec:collapse-problem}

A self-supervised encoder trained to predict its own future
representations faces a degenerate solution. If the encoder maps
every input to the same constant vector $\mathbf{c}$, then every
predicted representation equals every target representation --
the prediction loss is zero, and the task is trivially solved. But
the representation carries no information: every state looks
identical to the encoder. This is \emph{representational collapse},
and it is not a rare failure mode. It is the global minimum of any
naive prediction loss. Without countermeasures, gradient descent
will find it.

\smallskip
The reason is structural. The prediction loss is typically a
similarity measure -- cosine similarity or squared distance --
between the predicted and target representations. The loss is
minimized when the two match perfectly. The easiest way to achieve
a perfect match is for both representations to be constant: if the
encoder always outputs the same vector, the prediction is always
correct regardless of the input. Unlike classification losses,
which assign diverse targets to different classes, a self-prediction
objective has a single goal -- match yourself -- and nothing in the
loss penalizes constant outputs. Any self-supervised method built
on prediction must actively prevent this outcome.
\Cref{fig:collapse} illustrates the contrast: useful
representations preserve meaningful distinctions between game
states, while collapsed representations map every state to the same
point.

\input{figures/collapse}

\smallskip
Contrastive methods avoid collapse by construction. Methods such as
SimCLR~\citep{chen2020simple} and CURL~\citep{srinivas2020curl}
include negative pairs in the loss: the representation of state
$s_t$ must be similar to a positive example (an augmented view of
$s_t$) but dissimilar to negative examples (representations of
other states in the batch). If the encoder collapsed to a constant,
all representations would be identical and the negative pairs
would score maximally -- the loss would be high. The negatives act
as a repulsive force that prevents all representations from
converging to a single point. But contrastive methods require
large batches of negatives to work well. SimCLR uses batches of
4{,}096 or more; RL typically uses replay batches of 32. This
mismatch is a key reason why non-contrastive methods are preferred
in RL -- a practical constraint, not just an aesthetic preference.

\smallskip
BYOL~\citep{grill2020bootstrap} demonstrated that collapse can be
prevented without negative pairs using two mechanisms together. The
first is an \emph{exponential moving average (EMA) target encoder}.
Instead of using the same encoder for both prediction and target
generation, BYOL maintains a separate target encoder whose
parameters are a slowly updated average of the online encoder:
\begin{equation}
\label{eq:ema-update}
\boldsymbol{\theta}_{\text{target}}
  \leftarrow \tau\, \boldsymbol{\theta}_{\text{target}}
  + (1 - \tau)\, \boldsymbol{\theta}_{\text{online}},
\end{equation}
where $\tau$ is close to 1 (typically 0.99). At each training step,
the target encoder retains 99\% of its previous parameters and
absorbs 1\% from the online encoder. Over hundreds of steps, the
target gradually follows the online encoder, creating a temporal
lag: the target reflects what the online encoder looked like in the
recent past, not what it looks like now.

\smallskip
This temporal lag provides a restoring force against collapse. If
the online encoder improves, the target slowly catches up,
providing progressively better prediction targets. If the online
encoder tries to collapse, the target still reflects its
pre-collapse state -- diverse representations from the recent past
-- so the prediction loss increases rather than decreases. The
collapse is self-correcting because the target does not collapse
simultaneously.

\smallskip
The second mechanism is an \emph{asymmetric predictor head}: a
small MLP that exists only on the online side, mapping the online
representation to the prediction space before comparison with the
target. The target encoder has no predictor. This asymmetry breaks
the symmetry that enables collapse. The predictor must learn a
non-trivial mapping from the online representation to the target
representation. If the encoder collapses, the predictor receives
the same constant input for every example and cannot learn a useful
mapping -- its gradient pushes the encoder back toward diverse
representations. Removing either mechanism causes collapse: the EMA
alone is not sufficient, and the predictor alone is not sufficient.
Both together are required.

\smallskip
The EMA target encoder is the same design pattern the reader
already knows from \cref{subsec:target-networks}: the DQN target
network is a frozen copy that provides stable Q-value targets to
prevent oscillation in Q-learning. The EMA target encoder serves
an analogous role for representation learning: a slow-moving copy
that provides stable representation targets to prevent collapse.
The difference is the update mechanism: DQN uses hard updates
(copy parameters every $C$ steps), while the EMA is a soft update
(blend parameters at every step), avoiding the periodic
discontinuities of hard copying. In SPR, both coexist as separate
components -- the target network provides Q-value targets, and the
EMA target encoder provides representation targets -- sharing the
same design pattern of stability through slow change.

\smallskip
With these mechanisms in place, a prediction-based self-supervised
objective can learn meaningful representations without collapse.
The next section describes SPR, which applies this approach to
predict the agent's own future latent states.

\subsection{Self-Predictive Representations}
\label{subsec:spr}

SPR~\citep{schwarzer2021dataefficient} trains the encoder to
predict its own future representations -- not future pixels, not
future rewards, but future latent states. The idea follows directly
from the previous two subsections: RL has a natural pretext task
(temporal prediction), predicting in latent space avoids the cost of
pixel reconstruction, and the EMA target encoder plus asymmetric
predictor prevent collapse. SPR adds one further insight:
\emph{multi-step} prediction -- predicting $K$ steps ahead rather
than just one -- produces richer representations because the
encoder must capture dynamics that unfold over multiple time steps.
By default, SPR uses $K = 5$.

\smallskip
SPR adds several components on top of the base agent's
architecture. The \emph{online encoder} is the standard CNN
encoder from DQN (the Nature CNN described in
\cref{subsec:conv-encoder}). It takes the observation $s_t$ and
produces a spatial feature map
$\mathbf{y}_t = f_{\text{online}}(s_t)$. This is the same
encoder that feeds into the Q-head; the SPR components share it
rather than using a separate network.

\smallskip
The \emph{EMA target encoder} is a slow-moving copy of the online
encoder, updated via the EMA rule from \cref{eq:ema-update}. It
takes future observations $s_{t+k}$ and produces target
representations
$\mathbf{y}^{\text{tgt}}_{t+k} = f_{\text{target}}(s_{t+k})$.
Gradients do not flow through this encoder -- it provides stable
targets for the online encoder to predict, using the same
anti-collapse mechanism described in \cref{subsec:collapse-problem}.

\smallskip
The \emph{transition model} $h$ is an action-conditioned network
that predicts the next latent state given the current latent and
the action taken:
$\hat{\mathbf{y}}_{t+1} = h(\mathbf{y}_t,\, a_t)$. This is the
component that captures dynamics: it learns how the latent
representation should change when the agent takes a specific
action. The transition model is applied iteratively for multi-step
prediction -- each step's output becomes the next step's input.

\smallskip
Two additional heads complete the architecture. A \emph{projection
head} $g$ maps spatial feature maps to a lower-dimensional vector
suitable for comparison. It is applied to both predicted and target
representations. A \emph{prediction head} $q$ -- the asymmetric
predictor from BYOL -- is applied only to the online (predicted)
side, providing the second anti-collapse mechanism from
\cref{subsec:collapse-problem}. The full comparison for a single
prediction step is between
$q(g(\hat{\mathbf{y}}_{t+k}))$ on the online side and
$g(\mathbf{y}^{\text{tgt}}_{t+k})$ on the target side.
\Cref{fig:spr-architecture} shows the forward prediction loop, and
\cref{fig:spr-loss} shows how the online and target paths are
compared.

\input{figures/spr-architecture}
\input{figures/spr-loss}

\smallskip
SPR's training operates on sequences sampled from the replay
buffer. Given a transition sequence
$(s_t, a_t, s_{t+1}, a_{t+1}, \ldots, s_{t+K})$, the procedure
is as follows. First, the current state is encoded with the
online encoder:
$\mathbf{y}_t = f_{\text{online}}(s_t)$. Then, for each of the
next $K$ steps, the transition model predicts the next latent from
the previous \emph{prediction} (not from a fresh encoding):
$\hat{\mathbf{y}}_{t+k} = h(\hat{\mathbf{y}}_{t+k-1},\, a_{t+k-1})$.
Meanwhile, the actual future state $s_{t+k}$ is encoded with the
target encoder to produce the prediction target
$\mathbf{y}^{\text{tgt}}_{t+k}$. The SPR loss is the negative mean
cosine similarity across all $K$ steps:
\begin{equation}
\label{eq:spr-loss}
\mathcal{L}_{\text{SPR}}
  = -\frac{1}{K} \sum_{k=1}^{K}
    \frac{q(g(\hat{\mathbf{y}}_{t+k})) \cdot
          g(\mathbf{y}^{\text{tgt}}_{t+k})}
         {\|q(g(\hat{\mathbf{y}}_{t+k}))\|\;
          \|g(\mathbf{y}^{\text{tgt}}_{t+k})\|}.
\end{equation}
Gradients flow through the online encoder, transition model,
projection head, and prediction head, but not through the target
encoder.

\smallskip
The chained predictions are a deliberate design choice, not a
shortcut. Because each step uses the \emph{previous prediction}
rather than a fresh encoding, errors compound: a poor prediction at
step~1 produces a flawed input for step~2, which in turn degrades
step~3, and so on. This compounding pressure forces the encoder to
produce representations that the transition model can propagate
accurately over multiple steps. The encoder must represent
information that is \emph{predictable over time} -- stable features
such as object positions and velocities, rather than transient
noise like exact pixel patterns. The $K = 5$ horizon is enough to
capture short-term dynamics (ball trajectories, paddle movements)
without requiring long-horizon planning.

\smallskip
The encoder thus receives gradients from two losses: the Q-learning
loss, which signals what actions are valuable, and the SPR loss,
which signals how states evolve under actions. The SPR loss
encourages the encoder to represent object positions, velocities,
and action effects -- information that is useful for Q-learning but
that the Q-loss alone may not explicitly require, especially early
in training when Q-targets are noisy. The result is richer, more
informative representations that support faster and more stable
Q-learning. The encoder captures the environment's dynamics, not
just whatever happens to correlate with current Q-targets.

\smallskip
SPR shapes representations but does not change the agent's
exploration strategy, replay sampling, or learning algorithm. The
agent still uses $\varepsilon$-greedy exploration and uniform
replay. SPR does not add domain knowledge: the transition model is
a generic convolutional network, and the representations it
encourages are discovered from data. Its benefits may therefore
depend on the base agent. A capable base agent such as Rainbow
provides richer replay data (through prioritized sampling and
$n$-step returns) and more effective exploration (through noisy
networks), which could make SPR's predictions more informative. A
simple DQN may provide weaker data for SPR to learn from.

\smallskip
This is precisely the question that motivates the experiments in
this work. SPR was originally evaluated on top of
Rainbow~\citep{hessel2018rainbow}, which includes six extensions
beyond DQN. The reported improvements on Atari-100K were
substantial, but the base agent was already sophisticated. The
gains could come from SPR's representation learning alone, from
the interaction between SPR and Rainbow's components, or from a
combination of both. By testing SPR on a simple DQN -- without
Rainbow's extensions -- the contribution of representation learning
can be isolated from the contribution of the base agent. Combined
with data augmentation as a second experimental factor, this
produces a controlled study of when self-supervised representations
help and what determines their usefulness.

\section{Related Work and Positioning}
\label{sec:related-work}

The previous two sections introduced the methods this work studies:
data augmentation and self-supervised representation learning. But
these are not the only approaches to data-efficient RL. Three other
families of methods have demonstrated strong results on Atari-100K:
world models, planning methods, and training efficiency
improvements. This section briefly surveys each family, then
positions the contribution of this work relative to the field. The
survey describes what each approach does and how it relates to our
work, not how it works internally -- the reader needs context, not
a second background chapter.

\subsection{World Models}
\label{subsec:world-models}

A world model learns to predict what happens next: given the
current state and action, predict the next state and reward. Once
learned, the model can generate imagined experience -- the agent
can simulate trajectories without interacting with the real
environment. This addresses data efficiency directly: if real
interactions are expensive, generate synthetic ones from the model.
SimPLe~\citep{kaiser2020model} demonstrated this approach on Atari
by learning a video prediction model from real frames, then
training the agent on a mixture of real and model-generated
rollouts. Pixel-level prediction is expensive, however, and model
errors compound over long imagined horizons, making the generated
frames increasingly unreliable.

\smallskip
Dreamer~\citep{hafner2020dream} and
DreamerV2~\citep{hafner2021mastering} address this limitation by
learning a world model in \emph{latent space}. The model predicts
future latent states rather than future pixels, and the policy is
trained entirely in the model's imagination. This is more efficient
than pixel prediction and represents a fundamentally different
paradigm: the agent's decision-making is model-based, while DQN
(even with SPR) remains model-free.

\smallskip
World models and SPR share a structural similarity: both learn to
predict future states. But they differ in what the predictions are
used for. World models use predictions to generate experience for
training or planning -- the model is a core component of the
agent's decision-making. SPR uses predictions only as a training
signal to shape the encoder's representations; the transition model
is discarded at test time. SPR is therefore simpler -- it does not
need accurate long-horizon predictions or reward prediction -- but
it cannot leverage model-generated data.

\subsection{Planning Methods}
\label{subsec:planning-methods}

Planning methods combine a learned model with tree search.
MuZero~\citep{schrittwieser2020mastering} learns a latent model
optimized end-to-end for planning: at test time, the agent runs
Monte Carlo tree search (MCTS) in latent space, simulating possible
action sequences, evaluating them with the learned model, and
selecting the best action.
EfficientZero~\citep{ye2021mastering} adapted MuZero for the
data-efficient regime by adding self-supervised consistency
regularization, achieving superhuman performance on several
Atari-100K games -- the strongest results on the benchmark at the
time of publication.

\smallskip
Planning methods and SPR both learn predictive models, but they
differ in how those models are used. Planning methods use the
model at test time for action selection: MCTS requires many forward
passes per decision. SPR uses the model only at training time; the
agent is model-free at test time, with the same computational cost
as vanilla DQN (one forward pass per action). This work does not
compare against planning methods -- they address data efficiency
through a fundamentally different paradigm.

\subsection{Training Efficiency}
\label{subsec:training-efficiency}

A third family of methods improves data efficiency not by learning
a model or adding auxiliary objectives, but by training the
existing agent more effectively.
BBF~\citep{schwarzer2023bigger} scales the encoder from the Nature
CNN to a larger ResNet architecture and introduces periodic network
resets to combat \emph{plasticity loss} -- the phenomenon where
neural networks gradually lose the ability to learn new features as
training progresses. BBF also uses SPR as part of its training
objective, meaning its strong results include SPR's contribution on
top of the network scaling and training improvements.
SR-SPR~\citep{doro2023sampleefficient} extends SPR's prediction
target to include successor representation features, combining
temporal prediction with value-based prediction to push replay
ratios higher than previously possible.

\smallskip
These methods change how the agent trains; this work studies what
the representations learn. BBF's use of SPR means its results
cannot be attributed to SPR alone -- the same attribution problem
that motivates our study, at a different level. The two lines of
work are complementary: we isolate SPR's representation learning
contribution, while training efficiency work asks how to maximize
a fixed training budget.

\subsection{Positioning}
\label{subsec:positioning}

\Cref{fig:method-landscape} summarizes the landscape. Each family
addresses the limited-data problem through a different mechanism,
but none ask when representation learning works or what determines
whether the learned representations are useful.

\input{figures/method-landscape}

\smallskip
The methods surveyed above all target data-efficient RL, but none
ask \emph{when} self-supervised representation learning works or
what determines whether the learned representations are useful. The
original SPR paper~\citep{schwarzer2021dataefficient} evaluated SPR
exclusively on Rainbow. Follow-up work -- BBF, SR-SPR -- continued
to use Rainbow or enhanced architectures as the base agent. No
published work has tested SPR on vanilla DQN, which means the field
does not know whether SPR's representation learning is
independently effective or whether it requires a capable base agent
to produce useful representations.

\smallskip
Existing evaluations also compare methods by final performance
scores -- mean return, human-normalized score, IQM -- without
examining what the learned representations actually encode. No
prior work has compared representations across different base
agents using interpretability tools such as linear probing or
latent space visualization. Such a comparison can reveal whether
SPR produces qualitatively different representations depending on
the base agent, or whether the same representations are learned
but exploited differently by the downstream Q-head.

\smallskip
This work does not aim to beat published scores on Atari-100K.
World models, planning methods, and training efficiency
improvements have all achieved results that a vanilla DQN study
cannot match. The value lies in answering a question that
performance-focused work cannot: what determines whether
self-supervised representations are useful? By isolating SPR from
its base agent, combining it with data augmentation as a second
experimental factor, and probing the learned representations across
capability levels, this work provides controlled evidence about the
interaction between representation learning and agent capability.
The following chapter describes how we implement this study: the
agents, the experimental design, and the interpretability methods
used to probe the learned representations.
