\chapter{Methods}
\label{ch:methods}

\section{DQN Implementation}
\label{sec:dqn-implementation}

Our DQN agent follows \citet{mnih2015human}. The architecture is the
Nature CNN described in \cref{subsec:conv-encoder}, trained with
RMSProp (learning rate $2.5 \times 10^{-4}$), a batch size of 32,
discount $\gamma = 0.99$, and gradient clipping at norm 10 (if the
gradient's magnitude exceeds 10, it is scaled down to prevent
excessively large updates). The
replay buffer samples uniformly, and the target network is updated
via hard copy. Rewards are clipped to $[-1, +1]$. All of these
match the original paper; we make no changes to the base algorithm.

\smallskip
The 100K-step budget requires adapting several scale-dependent
parameters. \Cref{tab:dqn-100k} lists the values that change.

\begin{table}[htbp]
\centering
\begin{tabular}{lll}
\toprule
Parameter & Standard DQN & Our 100K setting \\
\midrule
Training frames       & 10M        & 400K \\
Replay capacity       & 1M         & 100K \\
Warmup                & 50K steps  & 2K steps \\
Target network update & 10K steps  & 2K steps \\
$\varepsilon$ decay   & 1M frames  & 200K frames \\
Sticky actions $p$    & 0.0        & 0.25 \\
\bottomrule
\end{tabular}
\caption{Atari-100K adaptations. All other hyperparameters
  follow \citet{mnih2015human} unchanged.}
\label{tab:dqn-100k}
\end{table}

The replay buffer is sized to match the generation budget -- the
agent produces at most 100K transitions, so a larger buffer would
never fill. The warmup is reduced from 50K to 2K steps because the
original warmup would consume half the training budget before
learning begins. Epsilon decay is compressed to cover the first
half of training, and the target network updates every 2{,}000 steps
to maintain roughly 50 updates across the run. Sticky actions are
enabled following the revised evaluation protocol of
\citet{machado2018revisiting}. These settings follow the
data-efficient configuration from
\citet{schwarzer2021dataefficient}.

\smallskip
The agent is implemented in PyTorch using Gymnasium for the
environment interface. All experimental conditions in the DQN
isolation study (\cref{subsec:dqn-isolation}) share this base
implementation; the only differences are whether augmentation
and SPR are enabled.

\section{Data Augmentation}
\label{sec:augmentation-method}

We apply the random shift augmentation described in
\cref{subsec:random-shift} at training time: batches sampled from
the replay buffer are augmented before each gradient step. The
buffer itself stores raw, unaugmented observations. Each time a
transition is replayed, it receives a fresh random shift -- the
network never sees the exact same input twice.

\smallskip
The current state $s$ and next state $s'$ in each transition receive
independent random shifts (different crop positions). Within each
observation, all four stacked frames share the same shift to preserve
inter-frame motion information. The operation pads each
$84 \times 84$ frame by 4 pixels using reflection padding and takes
a random $84 \times 84$ crop, implemented on the GPU via Kornia.

\smallskip
When augmentation is combined with SPR, the augmented observations
are used for both the Q-learning loss and the SPR prediction loss.
This changes one SPR hyperparameter: the EMA target encoder momentum
$\tau$ is set to 0.0 (direct copy) instead of 0.99, following
\citet{schwarzer2021dataefficient}, who found that augmentation
provides sufficient regularization on its own, making the slow EMA
update unnecessary.

\section{Self-Predictive Representations}
\label{sec:spr-architecture}

SPR adds an auxiliary self-predictive objective to the base agent,
as described conceptually in \cref{subsec:spr}. This section
specifies our implementation: the component architectures, loss
computation, and hyperparameters. SPR is integrated identically
into both DQN and Rainbow so that the capability comparison
(\cref{subsec:capability-comparison}) is controlled.

\smallskip
The SPR loss is the negative cosine similarity between the online
prediction and the target representation, averaged over valid
predictions (those that do not cross episode boundaries):
%
\begin{equation}
\label{eq:spr-loss}
\mathcal{L}_{\text{SPR}}
  = -\frac{1}{N}\sum_{k=1}^{K}\sum_{b=1}^{B}
    m_{k,b}\;\frac{\tilde{y}_{k,b} \cdot \bar{y}_{k,b}}
    {\lVert\tilde{y}_{k,b}\rVert\;\lVert\bar{y}_{k,b}\rVert}
\end{equation}
%
where $\tilde{y}$ is the online prediction, $\bar{y}$ is the target,
$m_{k,b}$ masks out predictions that cross episode boundaries, and
$N$ is the number of valid predictions. This loss is combined with
the TD objective as in \cref{eq:ssl-loss}, with $\lambda = 2.0$.

\begin{table}[htbp]
\centering
\begin{tabular}{llll}
\toprule
Component & Architecture & Input & Output \\
\midrule
Transition $h$
    & 2$\times$ Conv 3$\times$3 + BN + ReLU
    & 64$\times$7$\times$7
    & 64$\times$7$\times$7 \\
Projection $g$
    & Linear + ReLU
    & 3136
    & 512 \\
Prediction $q$
    & Linear
    & 512
    & 512 \\
\bottomrule
\end{tabular}
\caption{SPR component architectures. EMA target components
  ($\bar{f}_\theta$, $\bar{g}$) mirror their online counterparts.}
\label{tab:spr-components}
\end{table}

\indent A key design choice is that the transition model operates on
the spatial feature maps ($64 \times 7 \times 7$ after Conv3) rather
than the flattened FC output. This preserves spatial structure,
allowing the model to condition on actions via spatial broadcasting
-- the one-hot action vector (a vector with one entry set to~1 for the
selected action and all others set to~0) is replicated across all
spatial positions and concatenated along the channel dimension. The
transition model itself is a two-layer convolutional network with
batch normalization, keeping the same spatial resolution as its
input.

\smallskip
The EMA target encoder momentum $\tau$ varies with augmentation
status, as described in \cref{sec:augmentation-method}. The
prediction depth is $K = 5$ steps.
\Cref{tab:spr-hyperparams} summarizes all SPR hyperparameters;
all values follow the original data-efficient
settings~\citep{schwarzer2021dataefficient}.

\smallskip
\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Prediction depth $K$         & 5    \\
Loss weight $\lambda$        & 2.0  \\
Projection dimension         & 512  \\
EMA momentum $\tau$ (no aug) & 0.99 \\
EMA momentum $\tau$ (w/ aug) & 0.0  \\
Transition model channels    & 64   \\
\bottomrule
\end{tabular}
\caption{SPR hyperparameters. The EMA momentum varies with
  augmentation status (see text).}
\label{tab:spr-hyperparams}
\end{table}

\section{Rainbow Implementation}
\label{sec:rainbow-implementation}

Rainbow combines the six extensions described in
\cref{subsec:rainbow-dqn}: distributional values (C51), prioritized
replay, dueling architecture, $n$-step returns, noisy exploration,
and double DQN. Our implementation follows
\citet{hessel2018rainbow} with the same 100K adaptations applied to
the DQN base (\cref{tab:dqn-100k}). The optimizer switches to Adam
(learning rate $6.25 \times 10^{-5}$, $\varepsilon = 1.5 \times
10^{-4}$) following the original Rainbow
settings. \Cref{tab:rainbow-hyperparams} lists the extension-specific
hyperparameters.

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
\multicolumn{2}{l}{\textit{C51 (distributional)}} \\
Atoms                   & 51 \\
$V_{\min}$, $V_{\max}$  & $-10$, $+10$ \\
\midrule
\multicolumn{2}{l}{\textit{Prioritized replay}} \\
Priority $\alpha$       & 0.5 \\
IS correction $\beta$   & 0.4 $\to$ 1.0 \\
\midrule
$n$-step returns        & $n = 3$ \\
Noisy nets              & Factorised Gaussian \\
Optimizer               & Adam, lr $= 6.25 \times 10^{-5}$ \\
\bottomrule
\end{tabular}
\caption{Rainbow hyperparameters. Dueling and double DQN are
  architectural choices with no additional hyperparameters.}
\label{tab:rainbow-hyperparams}
\end{table}

The dueling architecture splits the final layer into separate value
and advantage streams, and double DQN decouples action selection
from evaluation in the target computation. Neither introduces
additional hyperparameters. With noisy networks enabled, epsilon-greedy exploration is disabled ($\varepsilon = 0$) -- the network
learns its own exploration through parametric noise.

\section{Experiment Design}
\label{sec:experiment-design}

The experiments are organized in two parts. The first part is a
DQN isolation study that tests data augmentation and self-predictive
representations as independent factors on a simple base agent. The
second part is a capability comparison that tests SPR on Rainbow to
provide the contrasting case where representation learning is known
to produce large improvements. Together, the two parts answer whether
SPR's benefits require a capable base agent or transfer to a simpler
one.

\subsection{DQN Isolation Study}
\label{subsec:dqn-isolation}

The first part tests four conditions on vanilla DQN, each adding
one or both of the methods introduced in
\cref{sec:augmentation-method,sec:spr-architecture}.
\Cref{tab:dqn-conditions} lists the conditions and what each one
tests.

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Condition & Purpose \\
\midrule
DQN        & Baseline: Q-learning loss alone \\
DQN + Aug  & Contribution of augmentation \\
DQN + SPR  & Contribution of representation learning \\
DQN + Both & Interaction between the two \\
\bottomrule
\end{tabular}
\caption{DQN isolation study conditions.}
\label{tab:dqn-conditions}
\end{table}

All four conditions share the same encoder architecture, the same
hyperparameters, and the same training procedure
(\cref{sec:dqn-implementation}). The only differences are whether
random shift augmentation is applied to the sampled batches and
whether the SPR auxiliary loss is added to the training objective.
This controlled setup ensures that any performance differences
reflect the contribution of augmentation, SPR, or their combination
-- not confounding changes to the base agent.

\smallskip
Testing each method alone and in combination is necessary to
separate their individual contributions from their interaction.
Without the combined condition, we could not tell whether the two
methods produce gains beyond what either achieves alone.

\subsection{Capability Comparison}
\label{subsec:capability-comparison}

The second part adds two conditions using Rainbow
(\cref{sec:rainbow-implementation}) as the base agent: Rainbow
without SPR, and Rainbow with SPR integrated identically to the DQN
conditions -- same transition model, same loss weight, same
prediction depth. The comparison between DQN + SPR and Rainbow + SPR
holds the representation learning method constant and varies only
the base agent, directly testing whether SPR's effectiveness depends
on the agent's underlying capability.

\subsection{Game Selection}
\label{subsec:game-selection}

We select six games from the 26-game Atari-100K suite. The selection
is based on published Rainbow+SPR results in
\citet{schwarzer2021dataefficient}, which show that augmentation and
SPR interact differently across games. By choosing games that span
this range of interactions, we can test whether the same patterns
hold when SPR is applied to a simpler agent.
\Cref{tab:game-selection} lists the games and the interaction
pattern observed on Rainbow.

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Game & Published interaction (Rainbow + SPR) \\
\midrule
Boxing        & Both methods contribute independently \\
Crazy Climber & SPR dominant; augmentation adds little \\
Frostbite     & SPR dominant; sparse reward setting \\
Kangaroo      & Strong synergy; both needed \\
Road Runner   & SPR large lift; augmentation moderate \\
Up N Down     & Augmentation dominant; SPR depends on it \\
\bottomrule
\end{tabular}
\caption{Game selection. Interaction patterns are from
  Rainbow+SPR~\citep{schwarzer2021dataefficient}. A key question
  is whether these patterns transfer to vanilla DQN.}
\label{tab:game-selection}
\end{table}

If augmentation and SPR interacted identically across all games,
a single game would suffice. The diversity of published patterns
-- from independent contributions (Boxing) to strong synergy
(Kangaroo) to augmentation dominance (Up N Down) -- lets us test
whether the interaction depends on game-specific factors such as
reward density and visual complexity.

\subsection{Seeds and Statistical Reporting}
\label{subsec:seeds-stats}

Each condition is trained with three seeds per game (42, 13, 23),
chosen arbitrarily. Across six conditions and six games, this
produces 108 runs. We report per-game performance as the mean
return $\pm$ standard deviation across seeds. For aggregate
comparisons across games, we use the interquartile mean (IQM) with
stratified bootstrap 95\% confidence intervals, following
\citet{agarwal2021deep}.

\smallskip
Three seeds per condition is standard in the Atari-100K literature
but limits the precision of per-game comparisons. Our analysis
therefore focuses on patterns that are consistent across games
rather than on individual point estimates.

\subsection{Evaluation Protocol}
\label{subsec:eval-protocol}

Each agent trains for 100{,}000 environment steps (400{,}000 frames
with the standard frame skip of 4). Evaluation is decoupled from
training: after each run completes, saved checkpoints are re-loaded
and evaluated over 30 episodes. DQN conditions use a near-greedy
policy ($\varepsilon = 0.05$), following
\citet{mnih2015humanlevel}. Rainbow conditions use fully greedy
action selection ($\varepsilon = 0$), since NoisyNets provide
learned exploration during training and evaluation uses the
mean network weights. Sticky actions ($p = 0.25$) are enabled
during both training and evaluation. The reported score is the
mean undiscounted return across the 30 episodes at the final
checkpoint.

\smallskip
Checkpoints are saved every 10{,}000 environment steps (40{,}000
frames), producing 10 per run. Beyond providing the final
performance score, these checkpoints supply the frozen
representations used for the interpretability analyses in
\cref{sec:interpretability-methods} and the learning curves that
track how performance evolves over training.

\subsection{Compute}
\label{subsec:compute}

All runs execute on Google Colab Pro using NVIDIA A100 GPUs for
training and T4 GPUs for evaluation. A DQN run (100K steps) takes
approximately 15--20 minutes on an A100; Rainbow runs take
25--30 minutes due to distributional values, prioritized replay,
and noisy networks. Evaluation of a single run's 10 checkpoints
(300 episodes total) takes approximately 2--5 minutes on a T4,
depending on the game's episode length.

\section{Interpretability Methods}
\label{sec:interpretability-methods}

This section describes the methods we use to analyze what the
encoder has learned and how its representations vary across
experimental conditions.

\subsection{Linear Probing}
\label{subsec:linear-probing}

The evaluation metrics from \cref{subsec:metrics} measure how well the
agent plays, but not what its encoder has learned to represent. A high
score does not guarantee that specific game-state variables like
positions, scores, or lives~(\cref{subsec:game-state-access}) are
captured in the representation.

\smallskip
To evaluate what information the encoder captures, we apply linear
probing~(\cref{subsec:representations}). The encoder $f_\theta$ is
first trained end-to-end via reinforcement learning, then its weights
are frozen~(\cref{fig:linear-probe}). A single linear layer is trained
on the frozen representations to predict the game-state variables
labeled by AtariARI~(\cref{subsec:game-state-access}).

\input{figures/linear-probe}

\smallskip
This approach lets us compare representations across experimental
conditions (e.g., with and without SPR, or across agent
architectures) on a common basis: we hold the probing procedure
fixed and vary only the encoder that produced the representations.
Differences in probe accuracy reflect differences in what the
encoder has learned to represent, not differences in probe capacity.

\smallskip
Concretely, if DQN+SPR achieves higher probe accuracy than DQN for ball
position, that means SPR has helped the encoder capture ball position. If
both conditions achieve similar accuracy, SPR has not changed what the
encoder captures for that variable. We repeat this comparison for each
labeled variable and each experimental condition in \cref{ch:results}.

\subsection{Probe Design}
\label{subsec:probe-design}

Each probe is a single fully connected layer mapping from the
representation $h_t \in \mathbb{R}^d$ to a target
$y \in \mathbb{R}^k$, trained with mean squared error for
continuous targets and cross-entropy for categorical targets. We
train probes on states sampled from completed evaluation episodes,
using the same set of states across all conditions to ensure
comparability. During each episode, we record the encoder's output
$h_t$ and the corresponding RAM values at each step, producing paired
observations for probe training.

\smallskip
The probe targets are the game-state variables labeled by
AtariARI~(\cref{subsec:game-state-access}), including positions,
scores, and lives. For each target, we report probe accuracy
(classification) or $R^2$ (regression) on a held-out test set. The
specific targets vary by game and are defined in
\cref{sec:experiment-design}. \Cref{tab:probe-training} summarizes the probe training
configuration.

\smallskip
\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Probe architecture  & Single linear layer \\
Training states     & \textit{TBD} sampled from evaluation episodes \\
Train/test split    & \textit{TBD} \\
Optimizer           & Adam \\
Learning rate       & \textit{TBD} \\
Training epochs     & \textit{TBD} \\
Loss (continuous)   & Mean squared error \\
Loss (categorical)  & Cross-entropy \\
\bottomrule
\end{tabular}
\caption{Probe training configuration. Italicized values are
  placeholders to be finalized during experiment design.}
\label{tab:probe-training}
\end{table}

\subsection{Probe Attachment Points}
\label{subsec:probe-attachment}

We attach independent probes at multiple points in the encoder to
understand how information develops through the network
(\cref{fig:linear-probe-layers}). Each probe is trained separately
with its own linear layer. \Cref{tab:probe-attachments} lists the
attachment points and the corresponding representation dimensions.

\smallskip
\begin{table}[htbp]
\centering
\begin{tabular}{llll}
\toprule
Probe & Attachment point & Shape & Probe input dim \\
\midrule
A & After Conv3 & $64 \times 7 \times 7$ & 3136 \\
B & After FC    & 512                     & 512  \\
C & Final $h_t$ & 512                     & 512  \\
\bottomrule
\end{tabular}
\caption{Probe attachment points in the Nature CNN encoder.
  Labels correspond to \cref{fig:linear-probe-layers}.}
\label{tab:probe-attachments}
\end{table}

Probing at multiple depths reveals whether information is present
early (in spatial feature maps) or only emerges after compression
into the fully connected layers. This is informative for
understanding how SPR and data augmentation reshape the encoder's
internal representations, not just its final output.

\subsection{Latent Space Visualization}
\label{subsec:latent-viz}

While linear probes quantify what information a representation
contains, they do not reveal how that information is organized.
Two encoders may achieve similar probe accuracy yet structure their
representations very differently -- one separating game states into
tight clusters, the other spreading them across a diffuse manifold.
Dimensionality reduction techniques let us visualize this structure
directly.

We apply both t-SNE~\citep{vandermaaten2008visualizing} and
UMAP~\citep{mcinnes2018umap} to the encoder's representations,
projecting them from $\mathbb{R}^d$ to $\mathbb{R}^2$ for
visualization. Using both methods guards against artifacts: t-SNE
emphasizes local neighborhood structure and tends to reveal
well-separated clusters, while UMAP better preserves global
relationships and distances between clusters. Patterns that appear
consistently under both methods are more likely to reflect genuine
structure in the representation.

\smallskip
For each experimental condition and each probe attachment point
(\cref{tab:probe-attachments}), we collect representations from the
same evaluation episodes used for probing. Each point in the
resulting scatter plot corresponds to a single state $s_t$, colored
by a game-state property (e.g., object position, score) to reveal
whether the encoder organizes its representations around
task-relevant features. We generate side-by-side plots across the
four experimental conditions to compare how augmentation and SPR
reshape the latent space.

\smallskip
\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
t-SNE perplexity   & \textit{TBD} \\
t-SNE iterations   & \textit{TBD} \\
UMAP neighbors     & \textit{TBD} \\
UMAP min.\ distance & \textit{TBD} \\
Representations     & Same states as probe training set \\
Coloring            & Game-state properties (per-game) \\
\bottomrule
\end{tabular}
\caption{Dimensionality reduction parameters. Italicized values
  to be finalized during experiment design.}
\label{tab:viz-params}
\end{table}

Comparing visualizations across layers shows whether structure
emerges gradually through the network or appears abruptly after a
specific layer. Comparing across conditions reveals whether SPR or
augmentation produces more structured representations, complementing
the quantitative probe results with a qualitative view of
representation geometry.

\smallskip
Both methods are sensitive to hyperparameters and can produce
artifacts -- in t-SNE, cluster sizes and inter-cluster distances are
not meaningful~\citep{vandermaaten2008visualizing}. We report all
parameter values (\cref{tab:viz-params}), verify that conclusions
hold across a range of settings, and treat these visualizations as
complementary to the quantitative probe results.

\smallskip
\input{figures/latent-viz-guide}

\Cref{fig:viz-guide} illustrates the two extremes. In a structured
representation~(\cref{fig:viz-guide-good}), points with the same
game-state property form distinct clusters, indicating the encoder
organizes states around task-relevant features. In an unstructured
representation~(\cref{fig:viz-guide-bad}), the same coloring reveals
no grouping -- game-state properties are not separable in the
projected space. In practice, results fall between these extremes,
and the degree of clustering provides a qualitative measure of how
well the encoder captures the target property.

\subsection{Transition Model Analysis}
\label{subsec:transition-analysis}

SPR trains a transition model $g_\phi$ that predicts future latent
representations given the current representation and intervening
actions~(\cref{sec:spr-architecture}). During training, this model
drives the self-predictive loss that shapes the encoder's
representations. We analyze the transition model directly to
understand what kind of dynamics it has learned and how prediction
quality relates to representation quality.

\smallskip
For each SPR-enabled condition, we collect trajectories from
evaluation episodes and compute the cosine similarity between the
transition model's predicted representation
$\hat{h}_{t+k} = g_\phi(h_t, a_t, \ldots, a_{t+k-1})$ and the
encoder's actual representation $h_{t+k} = f_\theta(s_{t+k})$ at
each prediction horizon $k = 1, \ldots, K$, where $K$ is the SPR
prediction depth used during training. This produces a prediction
quality curve showing how faithfully the model anticipates future
representations as the horizon increases.

\smallskip
We examine this curve along two dimensions. First, the rate of
degradation over $k$ steps reveals how far into the future the
transition model remains reliable -- a slow decay suggests the model
has learned robust dynamics, while rapid decay indicates it relies
on short-range correlations. Second, we condition the analysis on
game-state properties (e.g., object proximity, score changes) to
test whether prediction quality varies with the complexity of the
underlying transition. States involving collisions or reward events
may be harder to predict than smooth trajectories.

\smallskip
This analysis applies only to the two experimental conditions where
SPR is enabled. By comparing transition model quality between the
SPR-only and augmentation+SPR conditions, we can assess whether
data augmentation improves or degrades the dynamics model. Combined
with the probe results from \cref{subsec:linear-probing}, this
reveals whether better predictions correspond to more informative
representations.

\subsection{Metric Correlation}
\label{subsec:metric-correlation}

If the capability comparison reveals that SPR helps Rainbow more
than DQN, the natural question is which Rainbow component enables
the difference. Rather than ablating all six components exhaustively,
we first screen candidates by correlating Rainbow training metrics
with representation quality across games.

\smallskip
For each game, we record per-condition metrics that reflect each
Rainbow extension's contribution: the distribution of replay
priorities (prioritized replay), the variance of $n$-step returns,
the entropy of the distributional value estimates -- a measure of
how spread out the distribution is, with low entropy indicating a
concentrated, confident estimate and high entropy indicating
uncertainty -- and the magnitude of noisy network parameters. We
then compute Spearman rank correlations -- which measure whether two
variables increase together monotonically, without assuming a linear
relationship -- between these per-game metrics and the per-game
improvement in probe accuracy that SPR produces on Rainbow versus
DQN. A strong correlation indicates that the extension's behavior
tracks representation quality across games, making it a candidate
for targeted ablation.

\subsection{Targeted Ablations}
\label{subsec:targeted-ablations}

The metric correlation identifies candidate components; targeted
ablations test them causally. For the one or two strongest
candidates from \cref{subsec:metric-correlation}, we add the
component to DQN + SPR and measure whether representation quality
improves. For example, if prioritized replay correlates most
strongly with probe accuracy, we train DQN + SPR + prioritized
replay on the same six games and compare probe results against
unmodified DQN + SPR.

\smallskip
Adding one component at a time isolates its contribution, avoiding
the confounds of the full Rainbow combination. This provides causal
evidence for which Rainbow property enables useful self-supervised
representations.

\section{Representation-Informed Modification}
\label{sec:repr-modification}

The interpretability analyses in \cref{sec:interpretability-methods}
are designed to reveal what the encoder represents under each
condition and what differs between DQN + SPR and Rainbow + SPR. If
that analysis identifies a specific bottleneck -- a missing property
that the capable agent provides but the simple agent lacks -- this
section describes how we test whether addressing it improves
performance.

\smallskip
The modification is one of three types, depending on what the probes
and ablations reveal: a change to SPR's training objective (e.g.,
modified loss weighting or prediction target), a change to the
training configuration (e.g., replay ratio or augmentation schedule),
or the addition of the single Rainbow component identified by the
targeted ablations in \cref{subsec:targeted-ablations}. In each
case, the modification targets the specific bottleneck identified
by the interpretability analysis, not a generic improvement.

\smallskip
The modified condition is evaluated on the same six games with the
same three seeds and the same evaluation protocol as all other
conditions (\cref{sec:experiment-design}). We compare against
unmodified DQN + SPR and measure both final performance and
representation quality (probe accuracy, latent space structure).
A successful modification should improve both -- a performance gain
without a corresponding change in representation quality would
suggest the improvement is unrelated to the bottleneck, while a
representation improvement without a performance gain would suggest
the bottleneck lies elsewhere in the agent.
